---
title: "Buoy T-S"
author: "Matt Dzaugis"
date: "7/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(lubridate)
library(broom)
```

## Read in Buoy data

```{r cars}
Buoys <- read_csv(here::here("Data/Buoy_data.csv"))
```


### Buoy based indicators

This markdown calculates the inshore stratification index used in the Developing Indicators of ecosystem change report. It also contains some exploratory analysis of salinity data.

```{r read_data}
Buoy_csv <- read_csv(here::here("Data/Buoy_data.csv"))
```

```{r basic_plot_funs}
yrday_plot <- function(type, avgInterval, depth, Bandwidth, variable){
    plot1 <- switch(avgInterval,
                    
                  "day" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% 
                    mutate(yrday = yday(Date),
                           yr = year(Date)) %>% 
                            ggplot() + 
                    geom_point(aes(yrday, Values, col = as.factor(yr))) + 
                    facet_wrap(~name, ncol = 2) +
                    scale_color_viridis_d() + 
                    theme_bw() + theme(panel.grid = element_blank()),
                  
                  "mon" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% group_by(name) %>% 
                            mutate(
                                   mon = month(Date),
                                   yr = year(Date)) %>% group_by(mon, yr, name) %>% 
                            summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>% 
                            mutate(Date = as.Date(paste(yr, mon, "01", sep = "-"))) %>% 
                            ggplot() + geom_line(aes(mon, Values, col = as.factor(yr))) + 
                    facet_wrap(~name, ncol = 2) +
                    scale_color_viridis_d() + 
                    theme_bw()+ theme(panel.grid = element_blank()),
                  
                  "yr" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% group_by(name) %>% 
                            mutate(
                                   mon = month(Date),
                                   yr = year(Date)) %>% group_by(yr, name) %>% 
                            summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>% 
                            ggplot() + geom_point(aes(name, Values, col = as.factor(yr))) +
                    scale_color_viridis_d() + 
                    theme_bw()+ theme(panel.grid = element_blank()),
                  
                    "ksmooth" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% 
                    group_by(name, Type, Depth, Variable) %>% nest() %>% 
                    mutate(Values = purrr::map(data, 
                                               ~ksmooth(.x$Date, .x$Values, 
                                                        kernel = "normal", bandwidth = Bandwidth)),
           Values = purrr::map(Values, ~data.frame(.x))) %>% 
    select(-data) %>% unnest(Values) %>% mutate(x = as.Date(x), 
                                                yrday = yday(x), yr = year(x),
                                                Values = y) %>% 
    ggplot() + geom_line(aes(yrday, Values, col = as.factor(yr))) + 
                    facet_wrap(~name, ncol = 2) +
                    scale_color_viridis_d() + 
      theme_bw() + theme(panel.grid = element_blank()))
                            
  return(plot1)
}

basic_plot <- function(type, avgInterval, depth, variable){
  
  plot1 <- switch(avgInterval,
                  "day" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% 
                            ggplot() + geom_line(aes(Date, Values)) + facet_wrap(~name, ncol = 2) + 
                    theme_bw() + theme(panel.grid = element_blank()),
                  
                  "mon" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% group_by(name) %>% 
                            mutate(
                                   mon = month(Date),
                                   yr = year(Date)) %>% group_by(mon, yr, name) %>% 
                            summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop") %>% 
                            mutate(Date = as.Date(paste(yr, mon, "01", sep = "-"))) %>% 
                            ggplot() + geom_line(aes(Date, Values)) + facet_wrap(~name, ncol = 2) + 
                    theme_bw() + theme(panel.grid = element_blank()),
                  
                  "yr" = Buoy_csv %>% 
                    filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% group_by(name) %>% 
                            mutate(
                                   mon = month(Date),
                                   yr = year(Date)) %>% group_by(yr, name) %>% 
                            summarise(Values = mean(Values, na.rm = TRUE), .groups = "drop")  %>% 
                            ggplot() + 
                    geom_line(aes(yr, Values)) + facet_wrap(~name, ncol = 2) + labs(x = "Year") + 
                    theme_bw() + theme(panel.grid = element_blank()))
                            
  return(plot1)
}


```

```{r smoothing_plot_funs}


sma_plot <- function(type, depth, Bandwidth, variable){
  
  Buoy_csv %>% filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% 
    mutate(Values = zoo::rollmean(Values, Bandwidth, fill = NA, align = "right")) %>% 
                            ggplot() + geom_line(aes(Date, Values)) + facet_wrap(~name, ncol = 2) + 
    theme_bw() + theme(panel.grid = element_blank())

}

triangularma_plot <- function(type, depth, Bandwidth, variable){
  
  Buoy_csv %>% filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% 
    mutate(Values = zoo::rollmean(zoo::rollmean(Values, Bandwidth, fill = NA, align = "right"), Bandwidth, fill = NA, align = "right")) %>% 
                            ggplot() + geom_line(aes(Date, Values)) + facet_wrap(~name, ncol = 2) + theme_bw() + theme(panel.grid = element_blank())

}

KernelSmooth_plot <- function(type, depth, Bandwidth, variable){
  
  Buoy_csv %>% filter(Variable == variable) %>% 
                            filter(Type == type, Depth == depth) %>% group_by(name, Type, Depth, Variable) %>% nest() %>% 
    mutate(Values = purrr::map(data, ~ksmooth(.x$Date, .x$Values, kernel = "normal", bandwidth = Bandwidth)),
           Values = purrr::map(Values, ~data.frame(.x))) %>% 
    select(-data) %>% unnest(Values) %>% mutate(Values = y) %>% 
    ggplot() + geom_line(aes(x,Values)) + facet_wrap(~name, ncol = 2) + theme_bw() + theme(panel.grid = element_blank())

}

```

### Extreme saltwater and freshwater events

The plots show the 1st and 99th percentiles of salinity, where the y axis shows the number 1% or 99% extreme events per year.

```{r fw_events}

fwsw <- Buoy_csv %>% 
    filter(name %in% c("Buoy B", "Buoy E", "Buoy F", "Buoy I"), Variable == "sal") %>% 
    filter(Type == "Anomaly") %>% 
    mutate(mon = month(Date),
           yr = year(Date)) %>% 
    group_by(name, Type, Depth, Variable) %>% 
    mutate(quantile = scales::percent(.99),
           firstq = scales::percent(.01),
           ninty5th = quantile(Values, .99, na.rm = TRUE),
           fifth = quantile(Values, .01, na.rm = TRUE)) %>% 
    mutate(SW = if_else(Values > ninty5th, 1, 0),
           FW = if_else(Values < fifth, 1, 0)) %>% 
    group_by(Type, Depth, Variable, yr) %>% 
    summarise(FW = sum(FW, na.rm = TRUE),
              SW = sum(SW, na.rm = TRUE),
              .groups = "drop") 

fwsw %>% ggplot() + geom_col(aes(yr, SW)) + facet_wrap(~Depth, ncol = 1) + theme_bw() + theme(panel.grid = element_blank())
fwsw %>% ggplot() + geom_col(aes(yr, FW)) + facet_wrap(~Depth, ncol = 1) + theme_bw() + theme(panel.grid = element_blank())

```


## T-S relationship

```{r T_S, echo=FALSE}

T_S <- Buoys %>% 
  mutate(yr = year(Date)) %>% filter(Variable %in% c("temp", "sal"), Type == "Anomaly") %>% 
  group_by(name, Depth, yr) %>% pivot_wider(names_from = Variable, values_from = Values) %>%
  na.omit() %>%
  add_tally() %>%
  filter(n > 50) %>%
  nest() %>% 
  mutate(mod = map(data, ~lm(temp ~ sal, data = .x)),
         tidied = map(mod, tidy)) %>%
  unnest(tidied) %>% filter(term == "sal")

T_S %>% filter(name %in% c("Buoy M", "Buoy N")) %>% ggplot() + geom_line(aes(yr, estimate, col = as.factor(Depth))) + facet_wrap(~name) + theme_bw() + labs(y = "Slope", x = "Year", color = "Depth") + geom_smooth(aes(yr, estimate, col = as.factor(Depth)), method = "lm", se = FALSE)

```
