---
title: "COCA data"
author: "Matt Dzaugis"
date: "10/21/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## GARFO Data

shared Laz
Research 
COCA-Conf

CFDERS - vol and val of sale - dealer data

Dealer data tied to where they land the fish (landing port)

one record per community per species per year  - sum together

Port Code, Port Name, Port State, Year, NESPP4, NESPP3, SPP_COMMON_NAME, SPPLNDLB (species landed lbs), SPPLIVLB (species live lbs - with shell), SPPVALUE, STATE_DNUM, DEALER_NUM, HULLNUM, count of unique dealer and count of unique vessel 

for each year, spp, place - sum unique dealer and vessels
Sum unique dealer and vessel by place w/o species in place

merge dealer id and vessel num - start with federal and if fed is missing take the state 

Keep uniqueness based on the level of aggregation - need separate counts based on the groupings

Dealer Num is federal
STATE_DNUM is the state dealer number

at least back to 1996 is reliable (CFDERS)

Lobster - community level - eliminate some communities with very rare occurrences of landings 



```{r coca_headers}

coca <- read_csv("/Volumes/shared/Research/COCA-conf/Communities_June2021/KMills_Community_Data_2013-2020.csv", n_max = 0)

write_csv(coca, "COCA_headers.csv")
```





