---
title: "Analysis of Indicator"
author: "Matt Dzaugis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
        collapsed: FALSE
    code_folding: hide
editor_options: 
  chunk_output_type: console
params:
  indicator:
    label: "Select an indicator to process"
    value: "fvcom_bt"
    input: select
    choices: ["fvcom_bt","fvcom_sst","oisst","fvcom_bs","fvcom_sss","mcc","stratification","nefsc_biomass","nefsc_abundance","menh_abundance","menh_biomass","nefsc_size_spectra_slope","menh_size_spectra_slope","cpr_FirstMode","cpr_SecondMode"]
  season:
    label: "Select a season to process"
    value: "all"
    input: select
    choices: ["all", "spring", "summer", "fall", "winter"]
  stat_area:
    label: "Select a statistical area to process"
    value: "511-513"
    input: select
    choices: ["511", "512", "513", "511-513"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
library(factoextra)
library(gmRi)
# Set theme  
theme_set(theme_bw())
```


```{r read_data}
# Indicators
allIndicators <- read_csv(here::here("Processed_Indicators/allIndicators.csv"))
allIndicators$Season <- factor(allIndicators$Season, levels= c("spring", "summer", "fall", "winter", "all"))
allIndicators$stat_area <- factor(allIndicators$stat_area, levels= c("511", "512", "513", "511-513"))
indicator <- data.frame("name" = unique(allIndicators$Indicator))
indicator <- indicator %>% 
  mutate(long_name = case_when(name == "fvcom_bt"~"FVCOM NECOFS bottom temperature",
                      name == "fvcom_sst"~"FVCOM NECOFS surface temperature",
                      name == "oisst"~"OISSTv2.1",
                      name == "fvcom_bs"~"FVCOM NECOFS bottom salinity",
                      name == "fvcom_sss"~"FVCOM NECOFS surface salinity",
                      name == "mcc"~"Maine Coastal Current Index",
                      name == "stratification"~"Stratification Index",
                      name == "nefsc_biomass"~"NEFSC lobster predator biomass",
                      name == "nefsc_abundance"~"NEFSC lobster predator abundance",
                      name == "menh_abundance"~"ME/NH inshore trawl lobster predator abundance",
                      name == "menh_biomass"~"ME/NH inshore trawl lobster predator biomass",
                      name == "nefsc_size_spectra_slope"~"NEFSC size-based predator index",
                      name == "menh_size_spectra_slope"~"ME/NH size-based predator index",
                      name == "cpr_FirstMode"~"Small zooplankton index",
                      name == "cpr_SecondMode"~"Calanus index"))
# Enter indicator, season, stat area of interest

indicator_sel <- params$indicator
season_sel <- params$season
stat_areas_sel <- params$stat_area
long_name <- indicator$long_name[indicator$name == indicator_sel]

df1 <- allIndicators %>% 
  filter(Indicator == indicator_sel)

seasons <- unique(df1$Season)
stat_areas <- unique(df1$stat_area)

if(season_sel %not in% seasons){
  season_sel <- "all"
  note1 <- "Selected season not available for this indicator. Returned result with 'all' seasons."
} else {
  note1 <- season_sel
}

if(stat_areas_sel %not in% stat_areas){
  stat_areas_sel <- "511-513"
  note2 <- "Selected stat area not available for this indicator. Returned result with stat areas '511-513'."
} else {
  note2 <- stat_areas_sel
}

df <- allIndicators %>% 
    filter(Indicator == indicator_sel,
           Season == season_sel,
           stat_area == stat_areas_sel) %>% 
    na.omit()
```


```{r functions}
plot_fun <- function(x){
  x %>% 
    ggplot() + 
    geom_line(aes(Year, Value, col = stat_area)) +
    theme_bw() +
    facet_wrap(~Season)
}

sumLms_fun <- function(x){
  lm1 <- lm(Value ~ Year, data = x)
  return(lm1)
}


acf_plot <- function(x){
    df <- x %>% 
      select(Year, Value) %>% 
    na.omit()
  acf(df)
}

pscore_fun <- function(x, nbreak){
  df <- x
  
  lm1 <- lm(Value ~ Year, data = df)
  pscore <- segmented::pscore.test(lm1, n.break = nbreak)
  return(pscore)
}

davies_fun <- function(x){
  df <- x
  
  lm1 <- lm(Value ~ Year, data = df)
  davies <- segmented::davies.test(lm1)
  return(davies)
}

slope_bp_fun <- function(x, Npsi){
  df <- x
  
  lm1 <- lm(Value ~ Year, data = df)
  
  output <- tryCatch(segmented::segmented(lm1, seg.Z = ~Year, npsi = Npsi),
                     warning = function(c) c,
                     error = function(c) c)
  
  if(output[1] == "No breakpoint estimated"){
    print(paste(output[1]), "Returning lm")
  }
  
  if(output[1] == "$ operator is invalid for atomic vectors"){
      print(paste(output[1], "Error occurs in segmented() with very large numbers and two breakpoints. Scaled provides valid results."))
      lm1 <- lm(scale(Value) ~ Year, data = df)
  
      output <-segmented::segmented(lm1, seg.Z = ~Year, npsi = Npsi)
  }
  
  return(output)
  
}

mean_bp_fun <- function(x, models){
    df <- x
  
  lm1 <- lm(Value ~ Year, data = df)
  temp_mcp1 <- mcp::mcp(models, data = df, par_x = "Year")
  return(temp_mcp1)
}

```


`r use_gmri_style_rmd(css_file = "gmri_rmarkdown.css")`

## Developing Indicators of Habitat and Ecosystem Change in the Gulf of Maine

This report includes the initial analysis of the ecosystem and habitat indicators. Linear regression was used to assess trends over the length of the time series. A trend breakpoint analysis and a mean breakpoint analysis were run on the indicators. A breakpoint found in the slope of the line indicates a change in the trend of the data. A breakpoint in the mean of the date may indicate a regime shift or change of the overall state of the system. These analyses were run seasonally within statistical areas, when available, and as yearly averages across the entire study domain. 

## Indicators

The following indicators are used in this report

* `r long_name`

The following season and stat area are used in this report

* `r note1`
* `r note2`

### `r long_name` {.tabset}

#### Time series plot

```{r plots}
plot_fun(df)
```

#### Linear models

```{r lms}
lms <- sumLms_fun(df)
plot(lms)
summary(lms)
acf_plot(df)
```

#### Slope breakpoint tests

The pscore test and davies tests indicate whether a breakpoint in the slope exists. The davies test is more sensitive to data that follow a more sinusoidal curve but less sensitive to linear changes. A significant p-value indicates there is a breakpoint in the data. A non-significant p-value indicates a breakpoint is not present.

```{r breakpoint_tests}
pscore_fun(df, 1)
pscore_fun(df, 2)
davies_fun(df)
```

#### Slope breakpoint (1)

```{r slope_segments_1bp, warning=TRUE}
# Enter the npsi (estimated number of breakpoints) - might need to run the code first and assess the breakpoint stats
bp <- slope_bp_fun(df, 1)
summary(bp)
plot(bp)
points(x = df$Year, y = df$Value)
```

#### Slope breakpoint (2)

```{r slope_segments_2bp, warning=TRUE}
# Enter the npsi (estimated number of breakpoints) - might need to run the code first and assess the breakpoint stats
bp <- slope_bp_fun(df, 2)
summary(bp)
plot(bp)
points(x = df$Year, y = df$Value)
```

#### Slope breakpoint (3)

```{r slope_segments_3bp, warning=TRUE}
# Enter the npsi (estimated number of breakpoints) - might need to run the code first and assess the breakpoint stats
bp <- slope_bp_fun(df, 3)
summary(bp)
plot(bp)
points(x = df$Year, y = df$Value)
```

#### Mean breakpoint (1)

```{r mean_bp1}
# enter a list of models for the mean breakpoint - a model of 1~1 indicates intercept, adding additional models incudes additional bp
model_list <- list(Value~1, 1~1)
mean_cp <- mean_bp_fun(df, model_list)
summary(mean_cp)
plot(mean_cp)
```

#### Mean breakpoint (2)

```{r mean_bp2}
# enter a list of models for the mean breakpoint - a model of 1~1 indicates intercept, adding additional models incudes additional bp
model_list <- list(Value~1, 1~1, 1~1)
mean_cp <- mean_bp_fun(df, model_list)
summary(mean_cp)
plot(mean_cp)
```


