---
title: "Indicators of Habitat and Ecosystem Change"
author: "Matt Dzaugis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
        collapsed: FALSE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
library(factoextra)
library(gmRi)
# Set theme  
theme_set(theme_bw())
```


`r use_gmri_style_rmd(css_file = "gmri_rmarkdown.css")`

## Developing Indicators of Habitat and Ecosystem Change in the Gulf of Maine

```{r read_data}
# Indicators
FVCOM <- read_csv(here::here("Indicators/FVCOM_stat_area_temps.csv")) %>% filter(Year >= 1980)
OISST <- read_csv(here::here("Indicators/OISST_stat_area_anoms.csv")) %>% filter(Date <= as.Date("2020-12-31"))
sal <- read_csv(here::here("Indicators/FVCOM_stat_area_sal.csv")) %>% filter(Year >= 1980)
maineCC <- read_csv(here::here("Indicators/mcc_pca_pc1_2.csv")) %>% 
  rename("Year" = yr, "Month" = mon)
species_index <- read_rds(here::here("Indicators/nmfs_trawl_lobster_predator_indices.rdata")) %>% 
  rename("Year" = year, "stat_area" = `stratum id`)
mcc_turnoff_subset <- read_csv(here::here("Indicators/mcc_turnoff_subset.csv"))
menh_size_spectra <- read_csv(here::here("Indicators/MENH_size_spectra.csv"))
nefsc_size_spectra <- read_csv(here::here("Indicators/Nefsc_size_spectra.csv"))
cprData <- read_csv(here::here("Indicators/cpr_focal_pca_timeseries_period_1961-20017.csv")) %>% 
  mutate(`First Mode` = `First Mode`*-1) %>%
  rename("Year" = year,
         "FirstMode" = `First Mode`,
         "SecondMode" = `Second Mode`) %>% 
  select(Year, FirstMode, SecondMode)
strat_index <- read_csv(here::here("Indicators/stratification_index.csv")) %>% 
  rename("Year" = yr)
MEDMR_trawl_survey <- read_csv(paste0(gmRi::box_path("Res_Data", "Maine_NH_Trawl"), "MaineDMR_Trawl_Survey_Catch_Data_2021-05-14.csv"))

stat_area_key <- tibble("stat_area" = c(511,512,513),
                            "area" = c(296, 1381, 851))
stat_area_key$stat_area <- as.factor(stat_area_key$stat_area)


DMR_strata_area <- tibble("Stratum" = c("1", "2", "3", "4"),
                          "1" = c(253.27, 214.22, 227.35, 225.65),
                          "2" = c(279.63, 191.23, 211.66, 263.49),
                          "3" = c(259.62, 262.90, 280.03, 183.69),
                          "4" = c(205.30, 206.12, 310.49, 170.72),
                          "5" = c(138.54, 220.49, 365.04, 196.11)) %>% 
  pivot_longer(cols = c("1", "2", "3", "4", "5"), names_to = "Region", values_to = "strata_area") %>% 
  group_by(Region) %>% 
  mutate(Stratum = as.numeric(Stratum),
         total = sum(strata_area),
         Region = as.numeric(Region)) %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum) %>% 
  summarise(strata_area = sum(strata_area),
            total = sum(total))

seasons <- function(mon){
  x <- case_when(mon %in% c(12,1,2) ~ "winter",
                 mon %in% c(3,4,5) ~ "spring",
                 mon %in% c(6,7,8) ~ "summer",
                 mon %in% c(9,10,11) ~ "fall")
  return(x)
}
```


### Temperature

Surface and bottom temperature have a widely documented relationship with lobster biology. Temperature anomalies were calculated using the baseline climatology of 1990-2020 for each grid cell. Since FVCOM data are in an unstructured grid, temperatures were interpolated to a regular 0.1 x 0.1 deg grid and averaged over the statistical areas. OISSTv2.1 data were averaged over the statistical areas. Seasonal and yearly averages were calculated. 

Sources: 

* FVCOM NECOFS Monthly Means [Thredds Link](http://www.smast.umassd.edu:8080/thredds/catalog/models/fvcom/NECOFS/Archive/Seaplan_33_Hindcast_v1/monthly_mean/catalog.html)
* OISSTv2.1 [Information Link](https://www.ncei.noaa.gov/products/optimum-interpolation-sst)

```{r Temperature}

yrOISST <- OISST %>% 
  mutate(Year = lubridate::year(Date)) %>% 
  group_by(Year, stat_area) %>% 
  summarise(Value_var = var(temp),
            Value = mean(temp),
            .groups = "drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Season = "all",
         Indicator = "oisst")

seasonOISST  <- OISST %>% 
  mutate(Year = lubridate::year(Date),
         Month = lubridate::month(Date),
         Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year, stat_area) %>% 
  summarise(Value_var = var(temp),
            Value = mean(temp),
            .groups = "drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Indicator = "oisst")

yrFVCOM_sst <- FVCOM %>% 
  group_by(Year, stat_area) %>% 
  summarise(Value = mean(sur_temp_anom),
            Value_var = var(sur_temp_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Season = "all",
         Indicator = "fvcom_sst")

yrFVCOM_bt <- FVCOM %>% 
  group_by(Year, stat_area) %>% 
  summarise(Value = mean(bot_temp_anom),
            Value_var = var(bot_temp_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Season = "all",
         Indicator = "fvcom_bt")

seasonFVCOM_sst <- FVCOM %>% 
  mutate(Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year, stat_area) %>% 
  summarise(Value = mean(sur_temp_anom),
            Value_var = var(sur_temp_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Indicator = "fvcom_sst")

seasonFVCOM_bt <- FVCOM %>% 
  mutate(Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year, stat_area) %>% 
  summarise(Value = mean(bot_temp_anom),
            Value_var = var(bot_temp_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Indicator = "fvcom_bt")

TempStat_area <- bind_rows(yrOISST, seasonOISST, yrFVCOM_sst, yrFVCOM_bt, seasonFVCOM_sst, seasonFVCOM_bt)

TempDomain <- left_join(TempStat_area, stat_area_key, by = "stat_area") %>%  
  group_by(Year, Season, Indicator) %>% 
  summarise(Value = weighted.mean(Value, w = area/sum(unique(area))), .groups = "drop") %>% 
  mutate(stat_area = "511-513")

Temperature <- bind_rows(TempDomain, TempStat_area)

write_csv(Temperature, here::here("Processed_Indicators/Temperature.csv"))

```


#### Salinity

Salinity has been shown to have a relationship with lobster metabolism (Jury et al. 1994), lobster spatial distribution (Tanaka and Chen 2015), and incidence of shell disease (Tanaka et al. 2017). Salinity anomalies were calculated using the baseline climatology of 1990-2020 for each grid cell. Since FVCOM data are in an unstructured grid, temperatures were interpolated to a regular 0.1 x 0.1 deg grid and averaged over the statistical areas. 

Source: FVCOM NECOFS Monthly Means [Thredds Link](http://www.smast.umassd.edu:8080/thredds/catalog/models/fvcom/NECOFS/Archive/Seaplan_33_Hindcast_v1/monthly_mean/catalog.html)
Methods: 

* Surface and bottom layer (1 m above bathymetry)
* Regridded to regular 0.1 deg grid
* Averaged over NOAA statistical area for each year
* Baseline climatology 1990-2017 (last year of data)


```{r salinity}

yrFVCOM_sss <- sal %>% 
  group_by(Year, stat_area) %>% 
  summarise(Value = mean(sur_sal_anom),
            Value_var = var(sur_sal_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Season = "all",
         Indicator = "fvcom_sss")

yrFVCOM_bs <- sal %>% 
  group_by(Year, stat_area) %>% 
  summarise(Value = mean(bot_sal_anom),
            Value_var = var(bot_sal_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Season = "all",
         Indicator = "fvcom_bs")

seasonFVCOM_sss <- sal %>% 
  mutate(Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year, stat_area) %>% 
  summarise(Value = mean(sur_sal_anom),
            Value_var = var(sur_sal_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Indicator = "fvcom_sss")

seasonFVCOM_bs <- sal %>% 
  mutate(Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year, stat_area) %>% 
  summarise(Value = mean(bot_sal_anom),
            Value_var = var(bot_sal_anom),
            .groups="drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         Indicator = "fvcom_bs")

SalStat_area <- bind_rows(yrFVCOM_sss, yrFVCOM_bs, seasonFVCOM_sss, seasonFVCOM_bs)

SalDomain <- left_join(SalStat_area, stat_area_key, by = "stat_area") %>%  
  group_by(Year, Season, Indicator) %>% 
  summarise(Value = weighted.mean(Value, w = area/sum(unique(area))), 
            Value_var = weighted.mean(Value_var, w = area/sum(unique(area))),
            .groups = "drop") %>% 
  mutate(stat_area = "511-513")

Salinity <- bind_rows(SalDomain, SalStat_area)
write_csv(Salinity, here::here("Processed_Indicators/Salinity.csv"))
```


#### Maine Coastal Current Index

A bounding box extending along the coast of Maine capturing the location of the turn off offshore of Penobscot Bay was used to extract current strength and direction information.

Source: FVCOM NECOFS Monthly Means [Thredds Link](http://www.smast.umassd.edu:8080/thredds/catalog/models/fvcom/NECOFS/Archive/Seaplan_33_Hindcast_v1/monthly_mean/catalog.html)
Methods: 

* Surface layer
* Crop to Maine Coastal Current interest area
* Regridded to regular 0.1 deg grid
* Averaged over NOAA statistical area for each year
* See MCC_index_report.Rmd for details


```{r MCC}

yrMCC <- maineCC %>% 
  group_by(Year) %>% 
  summarise(Value = mean(PC1)) %>% 
  mutate(Season = "all",
         stat_area = "511-513",
         Indicator = "mcc")

seasonFVCOM_mcc <- maineCC %>% 
  mutate(Season = seasons(Month),
         Year = if_else(Month == 12, Year+1, Year)) %>% 
  group_by(Season, Year) %>% 
  summarise(Value = mean(PC1),
            Value_var = var(PC1),
            .groups="drop") %>% 
  mutate(stat_area = "511-513",
         Indicator = "mcc")

MCC <- bind_rows(yrMCC, seasonFVCOM_mcc)

write_csv(MCC, here::here("Processed_Indicators/MCC.csv"))
```


#### Stratification Index

* Brunt-Vaisala frequency N = sqrt(-g/potential density * delta density/delta depth)
* the ocean stratification is quantified by the measured value of delta density/delta depth
* If the water is more stratified, the frequency is higher. If less stratified, frequency is lower

Source: NERACOOS buoys

```{r strat_index}
strat_index_yr <- strat_index %>% 
  group_by(Year) %>% 
  summarise(Value = mean(index, na.rm = TRUE),
            Value_var = var(index, na.rm = TRUE), .groups = "drop") %>% 
  mutate(Season = "all",
         stat_area = "511-513",
         Indicator = "stratification")

strat_index_season <- strat_index %>% 
  mutate(Season = seasons(mon),
         Year = if_else(mon == 12, Year+1, Year)) %>% 
  group_by(Season, Year) %>% 
  summarise(Value = mean(index, na.rm = TRUE),
            Value_var = var(index, na.rm = TRUE),
            .groups="drop") %>% 
  mutate(stat_area = "511-513",
         Indicator = "stratification")

stratification <- bind_rows(strat_index_yr, strat_index_season)
write_csv(stratification, here::here("Processed_Indicators/stratification.csv"))
```


#### Species Based Predator Index - NEFSC

Source: NOAA NE Fisheries Trawl Survey
Methods: 

* Filtered to 15 lobster predators (ASMFC 2020)
* Cropped to NOAA stat areas 511, 512, 513
* Stratified abundance: abundance / km2 multiplied by the area of the strata
* Stratified biomass: biomass / km2 multiplied by the area of the strata




```{r nefsc_pred_index}
# Choose index `stratified biomass (kg)` or `stratified abundance`

spp_combine <- species_index %>% 
  mutate(season = as.character(season),
         stat_area = as.character(stat_area),
         season = tolower(season),
         Season = if_else(season == "both", "all", season),
         stat_area = if_else(stat_area == "Strata 511-513", "511-513", stat_area)) %>% 
  dplyr::select(Year, stat_area, 
                "nefsc_biomass" = `stratified biomass (kg)`, 
                "nefsc_abundance" = `stratified abundance`,
                Season) %>% 
  pivot_longer(cols = c(nefsc_biomass, nefsc_abundance), names_to = "Indicator", values_to = "Value")

write_csv(spp_combine, here::here("Processed_Indicators/nefsc_spp_pred_index.csv"))
```


#### Species Based Predator Index - ME/NH inshore trawl

Source: ME/NH inshore trawl Survey
Methods: 

* Filtered to 15 lobster predators (ASMFC 2020)
* Inshore trawl regions and strata were grouped into areas that most closely overlapped NOAA stat areas 511, 512, 513
* Stratified abundance cpue: abundance / tow multiplied by the area of the strata
* Stratified biomass cpue: biomass / tow multiplied by the area of the strata


```{r MEDMR_trawl pred_index}
lob_predators <- c("Halibut Atlantic", "Wolffish Atlantic", "Skate Barndoor", "Sea Bass Black", "Cod Atlantic", "Flounder Fourspot", "Haddock", "Skate Little", "Sculpin Longhorn", "Pout Ocean", "Hake Atlantic Red", "Sea Raven", "Hake Silver (Whiting)", "Skate Smooth", "Dogfish Spiny", "Hake Spotted", "Skate Thorny", "Hake White", "Flounder Winter") 
menh_pred <- MEDMR_trawl_survey %>% filter(Common_Name %in% lob_predators)

MEDMR_preds_stat_area <- menh_pred %>% filter(Common_Name != "Dogfish Spiny") %>% 
  left_join(stat_area_trawl_key) %>%  
  group_by(Year, Season, stat_area, Stratum)%>%
  mutate(tows=n_distinct(Tow_Number))%>%
  group_by(Year, Season, tows, stat_area, Stratum) %>%
  summarise(weight = sum(Expanded_Weight_kg,na.rm=TRUE), 
            catch=sum(Expanded_Catch,na.rm=TRUE)) %>%
  mutate(weight_tow = weight/tows, 
         catch_tow = catch/tows) %>% 
  left_join(DMR_strata_area) %>% 
  mutate(stratified_wpue = weight_tow*(strata_area/total),
         stratified_cpue = catch_tow*(strata_area/total)) %>% 
  group_by(stat_area, Year, Season, total) %>% 
  summarise(menh_abundance = sum(stratified_cpue), 
            menh_biomass = sum(stratified_wpue),
            .groups = "drop") %>% 
  mutate(stat_area = as.factor(stat_area))

yrMENH_preds_stat_area <- MEDMR_preds_stat_area %>% 
  group_by(Year, stat_area) %>% 
  summarise(menh_biomass = mean(menh_biomass),
         menh_abundance = mean(menh_abundance),
         Season = "all") %>% 
  ungroup()

yrMENH_preds <- MEDMR_preds_stat_area %>% 
  group_by(Year) %>% 
  summarise(menh_biomass = weighted.mean(menh_biomass, w = total/sum(unique(total))),
         menh_abundance = weighted.mean(menh_abundance, w = total/sum(unique(total))),
         Season = "all",
         stat_area = "511-513") %>% 
  ungroup()

MENH_preds <- bind_rows(MEDMR_preds_stat_area, yrMENH_preds_stat_area, yrMENH_preds) %>% 
  pivot_longer(cols = c(menh_abundance, menh_biomass), names_to = "Indicator", values_to = "Value") %>% 
  mutate(Season = tolower(Season))

write_csv(MENH_preds, here::here("Processed_Indicators/menh_spp_pred_index.csv"))
```


#### Size Based Predator Index - NEFSC

Source: NOAA NE Fisheries Trawl Survey


```{r size_based_index}
nefsc_size_spectra_all <- nefsc_size_spectra %>% 
  mutate(season = tolower(season),
         Season = if_else(season == "both", "all", season),
         stat_area = "511-513") %>% 
  select(Year, stat_area, Season, "Value" = b) %>% 
  mutate(Indicator = "nefsc_size_spectra_slope")

write_csv(nefsc_size_spectra_all, here::here("Processed_Indicators/nefsc_size_spectra.csv"))
```

#### Size Based Predator Index - ME/NH Trawl Survey

Source: NOAA NE Fisheries Trawl Survey

```{r menh_size_based_index}
menh_size_spectra_com <- menh_size_spectra %>% 
  mutate(season = tolower(season),
         Season = if_else(season == "both", "all", season),
         stat_area = "511-513") %>% 
  select(Year, stat_area, Season, "Value" = b) %>% 
  mutate(Indicator = "menh_size_spectra_slope")

write_csv(menh_size_spectra_com, here::here("Processed_Indicators/menh_size_spectra.csv"))
```


#### CPR data

Source: Continuous Plankton Recording

* First Mode = smaller zooplankton species explains ~ 50% of variance
* Second Mode = Calanus explains ~ 26% of variance

```{r cprData}
cprData_all <- cprData %>% 
  mutate(Season = "all",
         stat_area = "511-513") %>% 
  rename("cpr_FirstMode" = FirstMode,
         "cpr_SecondMode" = SecondMode) %>% 
  pivot_longer(cols = c(cpr_FirstMode, cpr_SecondMode), names_to = "Indicator", values_to = "Value")

write_csv(cprData_all, here::here("Processed_Indicators/cpr.csv"))
```


```{r allIndicators}

allIndicators <- bind_rows(Temperature, Salinity, MCC, stratification, spp_combine, MENH_preds, nefsc_size_spectra_all, menh_size_spectra_com, cprData_all) %>% 
  select(-Value_var, -total)

write_csv(allIndicators, here::here("Processed_Indicators/allIndicators.csv"))
```


