---
title: "Lobster Data Analysis"
author: "Matt Dzaugis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
        collapsed: FALSE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(patchwork)
library(factoextra)
library(gmRi)
# Set theme  
theme_set(theme_bw())
```

`r use_gmri_style_rmd(css_file = "gmri_rmarkdown.css")`

```{r read_data, message=FALSE}
all_lob_data <- read_csv(here::here("Processed_Indicators/all_lob_data.csv"))
```


```{r functions, message=FALSE}
# Function to run a segmented breakpoint analysis within a map function
# tryCatch captures any potential errors or warning message associated with no bp estimated
bp_analysis <- function(x, npsi){
  mod <-  lm(lob_index ~ Year, data = x)
  o <- tryCatch(segmented::segmented(mod, seg.Z = ~Year, npsi = npsi),  # need to estimate bp
                error = function(cond){cond})
}
# pscore test helps to determine if a breakpoint is present
pscore_fun <- function(x, Name, season, stat_areas, sex = "M+F", nbreak){
  df <- x %>% 
    filter(name == Name,
           Season == season,
           stat_area == stat_areas,
           Sex == sex)
  
  lm1 <- lm(lob_index ~ Year, data = df)
  pscore <- segmented::pscore.test(lm1, n.break = nbreak)
  return(pscore)
}
# davies test helps to determine if a breakpoint is present
davies_fun <- function(x, Name, season, stat_areas, sex = "M+F"){
  df <- x %>% 
    filter(name == Name,
           Season == season,
           stat_area == stat_areas,
           Sex == sex)
  
  lm1 <- lm(lob_index ~ Year, data = df)
  davies <- segmented::davies.test(lm1)
  return(davies)
}
# find breakpoints in the slope of the data
slope_bp_fun <- function(x, Name, season, stat_areas, sex = "M+F", Npsi){
    df <- x %>% 
    filter(name == Name,
           Season == season,
           stat_area == stat_areas,
           Sex == sex)
  
  lm1 <- lm(lob_index ~ Year, data = df)
  
  segmented::segmented(lm1, seg.Z = ~Year, npsi = Npsi)
}
# find breakpoints in the mean of the data
mean_bp_fun <- function(x, Name, season, stat_areas, sex = "M+F", models){
    df <- x %>% 
    filter(name == Name,
           Season == season,
           stat_area == stat_areas,
           Sex == sex)
  
  temp_mcp1 <- mcp::mcp(models, data = df, par_x = "Year")
  return(temp_mcp1)
}
# plot the sparkline
ggsparkline <- function(x, Name, season = "all", stat_areas = "511-513", sex = "M+F"){
  x %>% 
    filter(name == Name,
           Season == season,
           stat_area == stat_areas,
           Sex == sex) %>% 
    ggplot() + 
    geom_line(aes(Year, lob_index), size = .1) +
    scale_x_continuous(limits = c(1970, 2020)) +
    theme(panel.background = element_blank(),
          panel.grid = element_blank(),
          axis.line = element_blank(),
          text = element_blank(),
          axis.ticks = element_blank(),
          panel.border = element_blank())
}
```

## Developing Indicators of Habitat and Ecosystem Change in the Gulf of Maine

```{r define_filters}
season <- "all" # "spring" "fall" "all"
stat_areas <- "511-513" # "511" "512" "513" "511-513"
sex <- "M+F" # "Mal" "Fem" "M+F"
```

### Lobster Data{.tabset}

This report contains the results of an analysis of individual lobster datasets including: linear regression, breakpoint regression, and mean breakpoint (regime shift) tests. The following season(s), statistical areas, and lobster sex were analyzed

* Season: `r season`
* Statistical area: `r stat_areas`
* Sex: `r sex`

The following lobster datasets are used in this report

* NEFSC trawl survey abundance
* NEFSC trawl survey biomass
* NEFSC index from the ASMFC
* ME/NH index from the ASMFC
* ME/NH inshore trawl survey abundance
* ME/NH inshore trawl survey biomass
* Maine yearly landings
* Ventless trap survey sublegal lobsters
* American Lobster settlement index (ALSI)


```{r lms, message=FALSE}
all_names <- unique(all_lob_data$name)

lms <- all_lob_data %>% 
  filter(Season == "all", stat_area == "511-513", Sex == "M+F") %>% 
  group_by(name) %>% 
  nest() %>% 
  mutate(mod = map(data, ~lm(lob_index ~ Year, data = .x)),
         coefs = map(mod, ~broom::tidy(.x) %>% filter(term == "Year") %>% 
                       select(estimate, p.value)),
         rsq = map(mod, ~broom::glance(.x) %>% 
                     select(r.squared)),
         Model = "linear model") %>% 
  select(-data, -mod) %>% 
  unnest(cols = c(coefs, rsq))
```

#### NEFSC trawl survey abundance

```{r NEFSC_trawl_abun, message=FALSE}
#ggsparkline(all_lob_data, Name = "nefsc_abundance")
#ggsave(here::here("Plots/nefsc_lob_abundance_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "nefsc_abundance",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)

filter(lms, name == "nefsc_abundance")

max_breaks <- round(length(df$Year)/25)

pscore <- pscore_fun(all_lob_data, "nefsc_abundance", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "nefsc_abundance", season, stat_areas, sex, Npsi = max_breaks)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "nefsc_abundance", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)

nefsc_abundance_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### NEFSC trawl survey biomass

```{r NEFSC_trawl_bio, message=FALSE}
#ggsparkline(all_lob_data, Name = "nefsc_biomass")
#ggsave(here::here("Plots/nefsc_lob_biomass_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "nefsc_biomass",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)

filter(lms, name == "nefsc_biomass")

pscore <- pscore_fun(all_lob_data, "nefsc_biomass", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "nefsc_biomass", season, stat_areas, sex, Npsi = max_breaks)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)

mean_cp <- mean_bp_fun(all_lob_data, "nefsc_biomass", season, stat_areas, sex, models = list(lob_index~1, 1~1, 1~1))
summary(mean_cp)
plot(mean_cp)

nefsc_biomass_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### NEFSC index from the ASMFC

```{r NEFSCindex, message=FALSE}
#ggsparkline(all_lob_data, Name = "NEFSCindex")
#ggsave(here::here("Plots/NEFSCindex_lob_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "NEFSCindex",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)

filter(lms, name == "NEFSCindex")

pscore <- pscore_fun(all_lob_data, "NEFSCindex", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "NEFSCindex", season, stat_areas, sex, Npsi = 2)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "NEFSCindex", season, stat_areas, sex, models = list(lob_index~1, 1~1, 1~1))
summary(mean_cp)
plot(mean_cp)

NEFSCindex_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### ME/NH index from the ASMFC

```{r MENHindex, message=FALSE}
#ggsparkline(all_lob_data, Name = "MEindex")
#ggsave(here::here("Plots/MEindex_lob_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "MEindex",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)

filter(lms, name == "MEindex")

pscore <- pscore_fun(all_lob_data, "MEindex", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "MEindex", season, stat_areas, sex, Npsi =2)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "MEindex", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)

MEindex_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### ME/NH inshore trawl survey abundance

```{r MENH_trawl_abun, message=FALSE}
#ggsparkline(all_lob_data, Name = "menh_abundance")
#ggsave(here::here("Plots/menh_lob_abundance_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "menh_abundance",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)

filter(lms, name == "menh_abundance")

pscore <- pscore_fun(all_lob_data, "menh_abundance", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "menh_abundance", season, stat_areas, sex, Npsi = 2)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "menh_abundance", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)

menh_abundance_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### ME/NH inshore trawl survey biomass

```{r MENH_trawl_bio, message=FALSE}
#ggsparkline(all_lob_data, Name = "menh_biomass")
#ggsave(here::here("Plots/menh_lob_biomass_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "menh_biomass",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)


filter(lms, name == "menh_biomass")

pscore <- pscore_fun(all_lob_data, "menh_biomass", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "menh_biomass", season, stat_areas, sex, Npsi = 2)


pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "menh_biomass", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)

menh_biomass_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 2)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2) %>% 
  unnest(c(breakpoint1, breakpoint2))
```

#### Maine yearly landings

```{r yearly_landings, message=FALSE}
#ggsparkline(all_lob_data, Name = "ME_landings")
#ggsave(here::here("Plots/ME_lob_landings_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "ME_landings",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)


filter(lms, name == "ME_landings")

pscore <- pscore_fun(all_lob_data, "ME_landings", season, stat_areas, sex, nbreak = 2)
bp <- slope_bp_fun(all_lob_data, "ME_landings", season, stat_areas, sex, Npsi = 3)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "ME_landings", season, stat_areas, sex, models = list(lob_index~1, 1~1, 1~1))
summary(mean_cp)
plot(mean_cp)

ME_landings_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 3)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]]),
         breakpoint2 = purrr::map(mod, ~.x$psi[[2,2]]),
         breakpoint3 = purrr::map(mod, ~.x$psi[[3,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1, breakpoint2, breakpoint3) %>% 
  unnest(c(breakpoint1, breakpoint2, breakpoint3))
```

#### Ventless trap survey sublegal lobsters

```{r sublegal_data, message=FALSE}
#ggsparkline(all_lob_data, Name = "sublegal_cpue")
#ggsave(here::here("Plots/sublegal_cpue_sparkline.png"), width = 1, height = 1, units = "cm")

df <- all_lob_data %>% 
  filter(name == "sublegal_cpue",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)


filter(lms, name == "sublegal_cpue")

### No significant pscore
pscore <- pscore_fun(all_lob_data, "sublegal_cpue", season, stat_areas, sex, nbreak = 1)
bp <- slope_bp_fun(all_lob_data, "sublegal_cpue", season, stat_areas, sex, Npsi = max_breaks)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "sublegal_cpue", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)


sublegal_cpue_bp <- df %>% 
  select(name, stat_area, Season, Sex) %>% 
  unique() %>% 
  mutate(breakpoint1 = NA)
```

#### American Lobster settlement index (ALSI)

```{r ALSI_data, message=FALSE}
#ggsparkline(all_lob_data, Name = "ALSI")
#ggsave(here::here("Plots/ALSI_sparkline.png"), width = 1, height = 1, units = "cm")


df <- all_lob_data %>% 
  filter(name == "ALSI",
         Season == season,
         stat_area == stat_areas,
         Sex == sex)
max_breaks <- round(length(df$Year)/25)


filter(lms, name == "ALSI")

pscore <- pscore_fun(all_lob_data, "ALSI", season, stat_areas, sex, nbreak = 1)


bp <- slope_bp_fun(all_lob_data, "ALSI", season, stat_areas, sex, Npsi = 1)

pscore
summary(bp)
plot(bp) 
points(x = df$Year, y = df$lob_index)


mean_cp <- mean_bp_fun(all_lob_data, "ALSI", season, stat_areas, sex, models = list(lob_index~1, 1~1))
summary(mean_cp)
plot(mean_cp)

ALSI_bp <- df %>% 
  nest(data = -c(stat_area, name, Sex, Season)) %>% 
  mutate(mod = map(data, ~bp_analysis(.x, 1)),
         breakpoint1 = purrr::map(mod, ~.x$psi[[1,2]])) %>% 
  select(name, stat_area, Season, Sex, breakpoint1) %>% 
  unnest(c(breakpoint1))
```


```{r bind_bp, message=FALSE}
lob_breakpoints <- bind_rows(nefsc_biomass_bp, nefsc_abundance_bp, NEFSCindex_bp, MEindex_bp, ME_landings_bp, menh_abundance_bp, menh_biomass_bp, sublegal_cpue_bp, ALSI_bp)

#write_csv(lob_breakpoints, here::here("Processed_Indicators/lob_breakpoints.csv"))
```


