---
title: "Developing lobster data time series"
author: "Matt Dzaugis"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
        collapsed: FALSE
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(patchwork)
library(factoextra)
library(gmRi)
# Set theme  
theme_set(theme_bw())
```


`r use_gmri_style_rmd(css_file = "gmri_rmarkdown.css")`

## Developing lobster data time series

### Lobster Data {.tabset}


```{r biological_data}

ALSI_data <- read_csv(here::here("Biological_data/SettlementIndex.csv"))
ME_landings <- read_csv(here::here("Biological_data/ME_lob_landings_1950_2019.csv")) %>% 
  add_column("stat_area" = "511-513")
ASFMCindicies <- read_csv(here::here("Biological_data/GOMGBK_indices.csv")) %>% 
  rename("lob_index" = Index, "name" = Survey)
GOMindex <- ASFMCindicies %>% 
  filter(name %in% c("MeFQ2", "MeMQ2", "MeFQ4", "MeMQ4")) %>% 
  add_column("stat_area" = "511-513")
NEFSCindex <- ASFMCindicies %>% 
  filter(name %in% c("NefscFQ2", "NefscMQ2", "NefscFQ4", "NefscMQ4")) %>% 
  add_column("stat_area" = "511-513")
sublegal <- readxl::read_excel(here::here("Biological_data/RawData.xlsx")) %>% 
  rename("Year" = year, "Month" = month, "stat_area" = `stat area`)
MEDMR_trawl_survey <- read_csv(paste0(gmRi::box_path("Res_Data", "Maine_NH_Trawl"), "MaineDMR_Trawl_Survey_Catch_Data_2021-05-14.csv"))
MEDMR_trawl <- MEDMR_trawl_survey %>% 
  filter(Common_Name == "Lobster American")
NEFSC_meIndex <- read_csv(here::here("Biological_data/nmfs_trawl_lobster_indices.csv")) %>% 
  rename("Year" = est_year, "stat_area" = lobster_strata) %>% 
  mutate(stat_area = as.factor(stat_area))

```


#### ALSI index

Source: [American Lobster Settlement Index data portal](https://safis.accsp.org:8443/safis_prod/f?p=LSI:LOGIN)
Sites:

* Jonesport, Length: 2002-2018, stat area: 511
* Mt. Desert Island, Length: 2000-2018, stat area: 512
* Outer Penobscot Bay, Length: 2000-2018, stat area: 512
* Mid-coast, Length: 1989-2018, stat area: 513
* Casco Bay, Length: 2000-2018, stat area: 513
* York, Length: 2000-2018, stat area: 513

Methods:

* Sites grouped by NOAA stat area
* Time series cropped to shortest length
* Averaged by stat area

```{r ALSI_process}

statKey <- data.frame("Location" = c('Jonesport',
'Mt. Desert Island',
'Outer Pen Bay',
'Mid-coast',
'Casco Bay',
'York'), "stat_area" = c(511,512,512,513,513,513))

ALSI_stat_area <- ALSI_data %>% 
  left_join(statKey, by = "Location") %>% 
  filter(!is.na(stat_area),
         Year >= 2000) %>% 
  group_by(stat_area, Year) %>% 
  summarise(lob_index = mean(Yoy_density, na.rm=TRUE), .groups = "drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         name = "ALSI") %>% 
  na.omit()

ALSIyr <- ALSI_stat_area  %>% 
  group_by(Year, name) %>% 
  summarise(lob_index = mean(lob_index, na.rm=TRUE)) %>% 
  mutate(stat_area = "511-513")
  
ALSI <- bind_rows(ALSI_stat_area, ALSIyr)
ALSI$stat_area <- factor(ALSI$stat_area, levels = c("511", "512", "513", "511-513"))

ALSI$Season <- "all"
ALSI$Sex <- "M+F"

ALSI_plot <- ALSI %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "ALSI yoy density")

ALSI_plot

```

#### Sublegal index

Source: Ventless trap survey

Calculate stratified means

* Calculate catch per unit effort for each site for each year
* Multiply cpue by the depth strata area factor
* Group by stat area, sum the outputs 


```{r sublegal variance}

strata_area <- tibble("stat_area" = c(511,512,513),
                                 "1" = c(122, 566, 315),
                                 "2" = c(82, 395, 338),
                                 "3" = c(92, 420, 198)) %>% 
  pivot_longer(cols = c("1", "2", "3"), names_to = "depth stratum", values_to = "strata_area") %>% 
  group_by(stat_area) %>% 
  mutate(`depth stratum` = as.numeric(`depth stratum`),
         total = sum(strata_area)) 

# average number of lobsters per trap haul at each site
u <- sublegal  %>% 
  group_by(Year, `site ID`, `effort ID`, stat_area, `depth stratum`) %>% 
  summarise(n_lob = sum(`sample ID` != 0), .groups = "drop") %>% 
  group_by(Year, `site ID`, stat_area, `depth stratum`) %>% 
  summarise(n_lob_u = mean(n_lob), .groups = "drop")

# average number of lobsters per trap haul at each depth stratum within a stat area
v <- sublegal  %>% 
  group_by(stat_area, `depth stratum`, Year, `site ID`, `effort ID`) %>% 
  summarise(n_lob = sum(`sample ID` != 0), .groups = "drop") %>% 
  group_by(stat_area, `depth stratum`, Year) %>% 
  summarise(n_lob_v = mean(n_lob), .groups = "drop")

# choose the relevant v which corresponds to the depth strata that u is in ("stat_area", "depth stratum", "Year"). Each site will have one w value
w <- left_join(v, u, by = c("stat_area", "depth stratum", "Year")) %>% 
  mutate(w = (n_lob_v-n_lob_u)^2)

#sum of all w within the same depth strata in a stat area
x <- w %>% 
  group_by(`depth stratum`, stat_area, Year) %>% 
  summarise(x = sum(w), .groups = "drop")

#number of sites per depth stratum within a given stat area
y <- sublegal %>% 
  group_by(stat_area, `depth stratum`, Year) %>% 
  summarise(site_ID = unique(`site ID`)) %>% 
  summarise(y = n(), .groups = "drop")

#if done correctly, each stat area per year should have three z values, one for each depth stratum
z <- left_join(x, y, by = c("stat_area", "depth stratum", "Year")) %>% 
  mutate(z = x/(y-1))

# Calculate stat area variance 
sublegal_vari <- left_join(z, strata_area, by = c("stat_area", "depth stratum")) %>% 
  mutate(a = 1/(total^2),
         b = strata_area*(strata_area-y)*(z/y)) %>% 
  group_by(stat_area, Year, a) %>% 
  summarise(stat_sum = sum(b), .groups = "drop") %>% 
  mutate(vari = a*stat_sum,
         sd = sqrt(vari))

```


```{r sublegal lobs}

strata_area_factor <- tibble("stat_area" = c(511,512,513),
                                 "1" = c(0.412162162, 0.409847936, 0.370152761),
                                 "2" = c(0.277027027, 0.28602462, 0.397179788),
                                 "3" = c(0.310810811, 0.304127444, 0.23266745)) %>% 
  pivot_longer(cols = c("1", "2", "3"), names_to = "depth stratum", values_to = "strata_scale") %>% 
  mutate(`depth stratum` = as.numeric(`depth stratum`))

sublegal_cpue <- sublegal %>% 
  group_by(`trip ID`, `trip date`, Year, Month, `site ID`, 
           `depth stratum`, stat_area,
           `soak nights`, depth, `depth unit`, 
           `latitude (dd)`, `longitude (dd)`, `effort ID`) %>% 
  mutate(lob_count = sum(`sample ID` != 0), 
            lob_effort = lob_count/`soak nights`) %>% 
  group_by(`trip ID`, `trip date`, Year, Month, 
           `site ID`, `depth stratum`, stat_area,
           `soak nights`, depth, `depth unit`, 
           `latitude (dd)`, `longitude (dd)`) %>% 
  summarise(cpue = sum(lob_effort)/sum(!is.na(unique(`effort ID`))), .groups = "drop") %>% 
  left_join(strata_area_factor, by = c("stat_area", "depth stratum")) %>% 
  mutate(stratified_cpue = cpue*strata_scale,
         stat_area = as.factor(stat_area)) %>% 
  group_by(stat_area, Year) %>% 
  summarise(lob_index = sum(stratified_cpue), .groups = "drop") %>% 
  mutate(name = "sublegal_cpue")

sublegal_cpue_yr <- sublegal_cpue %>% 
  group_by(Year, name) %>% 
  summarise(lob_index = mean(lob_index, na.rm = TRUE)) %>% 
  mutate(stat_area = "511-513")
            
sublegal_cpue <- bind_rows(sublegal_cpue, sublegal_cpue_yr)
sublegal_cpue$stat_area <- factor(sublegal_cpue$stat_area, levels = c("511", "512", "513", "511-513"))

sublegal_cpue$Season <- "all"
sublegal_cpue$Sex <- "M+F"

sublegal_plot <- sublegal_cpue %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "Sublegal lobster cpue")

sublegal_var_plot <- sublegal_vari %>% 
  ggplot() +
  geom_line(aes(Year, vari, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "Sublegal lobster variance")


sublegal_plot


```


#### Gulf of Maine index

```{r MENHindex}

GOMindex <- GOMindex %>% 
  mutate(Season = if_else(Season == 2, "spring", "fall"),
         Season = as.factor(Season))

GOMindex_all_season <- GOMindex %>% 
  group_by(Sex, stat_area, Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Season = "all")

GOMindex_all_sex <- GOMindex %>% 
  group_by(Season, stat_area,Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Sex = "M+F")

GOMindex_all_season_sex <- GOMindex %>% 
  group_by(stat_area, Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Sex = "M+F",
         Season = "all")

GOMindex <- bind_rows(GOMindex, GOMindex_all_season, GOMindex_all_sex, GOMindex_all_season_sex)


GOMindex$name <- "MEindex"

GOMindex %>% 
  ggplot() +
  geom_line(aes(Year, lob_index)) +
  facet_wrap(~name) +
  labs(y = "Nefsc trawl lob abundance index")

```



#### ME-NH Trawl Survey

Source: ME-NH Trawl Survey

Calculate stratified means

* Calculate catch per unit effort for each site for each year
* Multiply cpue by the depth strata area factor
* Group by stat area, sum the outputs 

```{r ME trawl survey data}

stat_area_trawl_key <- tibble("stat_area" = c(511, 512, 512, 513, 513),
                              "Region" = c(5,4,3,2,1))

DMR_strata_area <- tibble("Stratum" = c("1", "2", "3", "4"),
                          "1" = c(253.27, 214.22, 227.35, 225.65),
                          "2" = c(279.63, 191.23, 211.66, 263.49),
                          "3" = c(259.62, 262.90, 280.03, 183.69),
                          "4" = c(205.30, 206.12, 310.49, 170.72),
                          "5" = c(138.54, 220.49, 365.04, 196.11)) %>% 
  pivot_longer(cols = c("1", "2", "3", "4", "5"), names_to = "Region", values_to = "strata_area") %>% 
  group_by(Region) %>% 
  mutate(Stratum = as.numeric(Stratum),
         total = sum(strata_area),
         Region = as.numeric(Region)) %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum) %>% 
  summarise(strata_area = sum(strata_area),
            total = sum(total))

# average number of lobsters per tow 
u <- MEDMR_trawl %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(Season, Year, Tow_Number, stat_area, Stratum) %>% 
  summarise(n_lob_u = sum(Expanded_Weight_kg, na.rm = TRUE), .groups = "drop")

# average number of lobsters per trap haul at each depth stratum within a stat area
v <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>%  
  group_by(stat_area, Stratum, Year, Season) %>% 
  summarise(n_lob_v = mean(Expanded_Weight_kg, na.rm = TRUE), .groups = "drop")

# choose the relevant v which corresponds to the depth strata that u is in ("stat_area", "depth stratum", "Year"). Each site will have one w value
w <- left_join(v, u, by = c("stat_area", "Stratum", "Year", "Season")) %>% 
  mutate(w = (n_lob_v-n_lob_u)^2)

#sum of all w within the same depth strata in a stat area
x <- w %>% 
  group_by(Stratum, stat_area, Year, Season) %>% 
  summarise(x = sum(w), .groups = "drop")

#number of sites per depth stratum within a given stat area
y <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum, Year, Season) %>% 
  summarise(Tow_number = unique(Tow_Number, na.rm = TRUE)) %>% 
  summarise(y = n(), .groups = "drop")

#if done correctly, each stat area per year should have three z values, one for each depth stratum
z <- left_join(x, y, by = c("stat_area", "Stratum", "Year", "Season")) %>% 
  mutate(z = x/(y-1))

# Calculate stat area variance 
MEDMR_vari <- left_join(z, DMR_strata_area, by = c("stat_area", "Stratum")) %>% 
  mutate(a = 1/(total^2),
         b = strata_area*(strata_area-y)*(z/y)) %>% 
  group_by(stat_area, Year, a, Season) %>% 
  summarise(stat_sum = sum(b), .groups = "drop") %>% 
  mutate(vari = a*stat_sum,
         sd = sqrt(vari))

medmr_vari_plot <- MEDMR_vari %>% 
  ggplot() + 
  geom_line(aes(Year, vari, col = as.factor(stat_area))) +
  facet_wrap(~Season)
```


```{r MEDMR_trawl cpue}

MEDMR_cpue <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>%  
  group_by(Year, Season, stat_area, Stratum)%>%
  mutate(tows=n_distinct(Tow_Number))%>%
  group_by(Year, Season, tows, stat_area, Stratum) %>%
  summarise(weight = sum(Expanded_Weight_kg,na.rm=TRUE), 
            catch=sum(Expanded_Catch,na.rm=TRUE))%>%
  mutate(weight_tow = weight/tows, 
         catch_tow = catch/tows) %>% 
  left_join(DMR_strata_area) %>% 
  mutate(stratified_wpue = weight_tow*(strata_area/total),
         stratified_cpue = catch_tow*(strata_area/total)) %>% 
  group_by(stat_area, Year, Season) %>% 
  summarise(menh_abundance = sum(stratified_cpue), 
            menh_biomass = sum(stratified_wpue),
            .groups = "drop") %>% 
  pivot_longer(cols = c(menh_abundance, menh_biomass), names_to = "name", values_to = "lob_index") %>% 
  mutate(stat_area = as.factor(stat_area))

MEDMR_cpue_yr <- MEDMR_cpue %>% 
  group_by(Year, name) %>% 
  summarise(lob_index = mean(lob_index, na.rm = TRUE)) %>% 
  mutate(Season = "all",
         stat_area = "511-513") %>% 
  mutate(stat_area = as.factor(stat_area))

MEDMR_cpue_season <- MEDMR_cpue %>% 
  group_by(Year, stat_area, name) %>% 
  summarise(lob_index = mean(lob_index, na.rm = TRUE)) %>% 
  mutate(Season = "all") %>% 
  mutate(stat_area = as.factor(stat_area))

MEDMR_cpue_stat <- MEDMR_cpue %>% 
  group_by(Year, Season, name) %>% 
  summarise(lob_index = mean(lob_index, na.rm = TRUE)) %>% 
  mutate(stat_area = "511-513") %>% 
  mutate(stat_area = as.factor(stat_area))

  
MENH_lob_index <- bind_rows(MEDMR_cpue, MEDMR_cpue_yr, MEDMR_cpue_stat, MEDMR_cpue_season) %>% 
  mutate(Season = tolower(Season))

MENH_lob_index$stat_area <- factor(MENH_lob_index$stat_area, levels = c("511", "512", "513", "511-513"))
MENH_lob_index$Sex <- "M+F"

MENH_lob_index %>% 
  filter(name == "menh_biomass") %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "ME-NH Trawl biomass") +
  facet_wrap(~Season)

MENH_lob_index %>% 
  filter(name == "menh_abundance") %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "ME-NH Trawl abundance") +
  facet_wrap(~Season)


```


#### ME yearly landings

Yearly Maine lobster landings in pounds from 1950-2020.

```{r yearly_landingd}

ME_pounds <- ME_landings %>% 
  mutate(name = "ME_landings",
         Season = "all",
         Sex = "M+F") %>% 
  select(Year, "lob_index" = Pounds, name, stat_area, Season, Sex) 

ME_pounds %>% 
  ggplot() +
  geom_line(aes(Year, lob_index)) +
  labs(y = "Maine lobster landings (pounds)")



```


#### Nefsc index

ASFMC lobster abundance index based on the NE fisheries trawl survey. Time spans 1978-2018 and indices are split into seasons and sex.

```{r Nefscindex}

NEFSCindex <- NEFSCindex %>% 
  mutate(Season = if_else(Season == 2, "spring", "fall"),
         Season = as.factor(Season))

NEFSC_all_season <- NEFSCindex %>% 
  group_by(Sex, stat_area, Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Season = "all")

NEFSC_all_sex <- NEFSCindex %>% 
  group_by(Season, stat_area, Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Sex = "M+F")

NEFSC_all_season_sex <- NEFSCindex %>% 
  group_by(stat_area, Year) %>% 
  summarise(lob_index = mean(lob_index)) %>% 
  mutate(Sex = "M+F",
         Season = "all")

NEFSCindex <- bind_rows(NEFSCindex, NEFSC_all_season, NEFSC_all_sex, NEFSC_all_season_sex)

NEFSCindex$name <- "NEFSCindex"

NEFSCindex %>% 
  ggplot() +
  geom_line(aes(Year, lob_index)) +
  facet_wrap(~name) +
  labs(y = "Nefsc trawl lob abundance index")


```


#### NMFS stat area 511-513

Lobster biomass and abundance for statistical areas 511-513

```{r nmfs_trawl}

NEFSC_meLobIndex <- NEFSC_meIndex %>% 
  rename("nefsc_biomass" = `stratified biomass (kg)`,
         "nefsc_abundance" = `stratified abundance`,
         "Season" = season) %>% 
  select(Year, Season, stat_area, nefsc_biomass, nefsc_abundance, `stratum area`) %>% 
  mutate(Season = tolower(Season),
         Season = if_else(Season == "both", "all", Season)) %>% 
  pivot_longer(cols = c(nefsc_biomass, nefsc_abundance), names_to = "name", values_to = "lob_index")
  

NEFSC_meIndex_yr <- NEFSC_meLobIndex %>% 
  mutate(total = sum(unique(`stratum area`))) %>% 
  group_by(Year, Season, name) %>% 
  summarise(lob_index = weighted.mean(lob_index, w = `stratum area`/total)) %>% 
  mutate(stat_area = "511-513")

NEFSC_lobIndex <- bind_rows(NEFSC_meLobIndex, NEFSC_meIndex_yr) %>% 
  select(-`stratum area`)
NEFSC_lobIndex$stat_area <- factor(NEFSC_lobIndex$stat_area, levels = c("511", "512", "513", "511-513"))

NEFSC_lobIndex$Sex <- "M+F"



NEFSC_lobIndex %>% 
  filter(name == "nefsc_biomass") %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "NEFSC Trawl biomass") +
  facet_wrap(~Season)

NEFSC_lobIndex %>% 
  filter(name == "nefsc_abundance") %>% 
  ggplot() +
  geom_line(aes(Year, lob_index, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "NEFSC Trawl abundance") +
  facet_wrap(~Season)

```


```{r all_lob_data}
all_lob_data <- bind_rows(NEFSC_lobIndex, NEFSCindex, GOMindex, ME_pounds, MENH_lob_index, sublegal_cpue, ALSI)

write_csv(all_lob_data, here::here("Processed_Indicators/all_lob_data.csv"))


```

