---
title: "Pred_index"
author: "Matt Dzaugis"
date: "11/29/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
library(rerddap)
library(foreign)
library(tidyverse)
library(sizeSpectra)
```


```{r data}
# Lobster Predators
lob_predators <- read_csv(here::here("Data/lobster_pred_list.csv"))

### for size spec
MENH_catch <- read_csv(here::here("Data/MaineDMR_Catch_at_Length_Data_2020-07-13.csv"), guess_max = 1e6)

coeffs <- read_csv(here::here("Data/listfishusingfishbase.csv")) %>% 
  janitor::clean_names() %>% dplyr::select(-x6)
```

### Lobster Predator Indices

The idea behind the lobster predator index is to use characteristics of the fish community that prey on lobster as a biological indicator for predicting patterns in lobster abundance/biomass/recruitment.

#### ME/NH catch summary

```{r weight, echo=FALSE}
# Species that are in grams
grams <- c(
  "Monkfish",
  "Lumpfish",
  "Squid Short-Finned",
  "Sculpin Shorthorn",
  "Crab Jonah",
  "Wrymouth",
  "Fourbeard Rockling",
  "Crab Atlantic Rock",
  "Blenny Snake",
  "Smelt Rainbow",
  "Sand Lance American",
  "Menhaden Atlantic",
  "Silverside Atlantic",
  "Alligatorfish"
)

#### FILTER FOR SEASON

catch <- MENH_catch %>% left_join(coeffs, by = c("Common_Name" = "common_name")) %>%
  filter(Unit_of_Length == "CM", 
         Common_Name != "Sea Urchins Green",
         !is.na(ln_a))  %>% 
  mutate(ln_weight = (ln_a + b * log(Length)),
         weight = if_else(Common_Name %in% grams, exp(ln_weight)/1000, exp(ln_weight)),
         freq_weight = weight*Frequency,
         a = exp(ln_a)) %>% 
  dplyr::select(-Sex) %>% 
  filter(Season == "Fall") #########33 LOOOK at season

catch %>% 
  ggplot( aes( y = fct_reorder(Common_Name, weight, .fun = mean, .desc = TRUE), x = weight)) + 
    geom_boxplot() + 
    labs(x = "Weight (kg)", y = "Common Name")

```

#### Catch cleanup and filter for predators

```{r cpue and preds}

stat_area_trawl_key <- tibble("stat_area" = c(511, 512, 512, 513, 513),
                              "Region" = c(1,2,3,4,5))

DMR_strata_area <- tibble("Stratum" = c("1", "2", "3", "4"),
                          "1" = c(253.27, 214.22, 227.35, 225.65),
                          "2" = c(279.63, 191.23, 211.66, 263.49),
                          "3" = c(259.62, 262.90, 280.03, 183.69),
                          "4" = c(205.30, 206.12, 310.49, 170.72),
                          "5" = c(138.54, 220.49, 365.04, 196.11)) %>% 
  pivot_longer(cols = c("1", "2", "3", "4", "5"), names_to = "Region", values_to = "strata_area") %>% 
  group_by(Region) %>% 
  mutate(Stratum = as.numeric(Stratum),
         total = sum(strata_area),
         Region = as.numeric(Region)) %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum) %>% 
  summarise(strata_area = sum(strata_area),
            total = sum(total))


cleanup_catch <- catch %>%
  left_join(stat_area_trawl_key) %>%  
  group_by(Year, stat_area, Stratum)%>%
  mutate(weight_g = weight*1000,
         ntows = n_distinct(Tow_Number)) %>% 
  ungroup() %>%
  rename("SpecCode" = Common_Name,
        "LngtClass" = Length,
        "Number" = Frequency,
        "LWa" = a,
        "LWb" = b,
        "bodyMass" = weight_g,
        "CPU_bio_per_hour" = freq_weight) %>% 
  left_join(DMR_strata_area) 

trawl_size_spec <- cleanup_catch  %>% 
  mutate(total_area = sum(unique(strata_area))) %>% 
  group_by(Year, stat_area, SpecCode, LngtClass, total_area, total) %>% # strata means
  summarise(
    Number = weighted.mean(Number, wt = strata_area/total), 
    LWa = unique(LWa),
    LWb = unique(LWb),
    bodyMass = unique(bodyMass), .groups = "drop") %>% 
  group_by(Year, SpecCode, LngtClass) %>%
  summarise(
    Number = weighted.mean(Number, wt = total/total_area), 
    LWa = unique(LWa),
    LWb = unique(LWb),
    bodyMass = unique(bodyMass), .groups = "drop") 

  #group_by(Year, Region, Stratum, Common_Name, Length) %>% 
  #add_tally() %>%
  #summarise(Frequency = sum(Frequency)/n, .groups = "drop") %>%
  #left_join(., MENH_area, by = c("Region", "Stratum")) %>% 
  #group_by(Year, Season, Common_Name, Length) %>%
  #summarise(Frequency = weighted.mean(Frequency, wt, na.rm=TRUE), .groups = "drop") %>%
  #left_join(., coeffs, by = "Common_Name")


# Set bodymass lower limit
mass_cutoff <- 400 #grams

# Filter for lower end of gear selectivity
trawl_size_spec <- filter(trawl_size_spec, bodyMass >= mass_cutoff)

lob_predators <- c("Halibut Atlantic", "Wolffish Atlantic", "Skate Barndoor", "Sea Bass Black", "Cod Atlantic", "Flounder Fourspot", "Haddock", "Skate Little", "Sculpin Longhorn", "Pout Ocean", "Hake Atlantic Red", "Sea Raven", "Hake Silver (Whiting)", "Skate Smooth", "Dogfish Spiny", "Hake Spotted", "Skate Thorny", "Hake White", "Flounder Winter") 

lob_size_spec <- trawl_size_spec %>% filter(SpecCode %in% lob_predators )

# Total number of fish
total_fish <- sum(lob_size_spec$Number)

# Biomass data
data_biomass <- lob_size_spec %>% 
  mutate(Biomass = Number * bodyMass)
```

#### Summary Plots

```{r look at data}
dataSum <- lob_size_spec %>% group_by(Year) %>% 
  dplyr::summarise(uniqLngClass = length(unique(LngtClass)),
                   uniqSpec = length(unique(SpecCode)),
                   minLngClass = min(LngtClass),
                   maxLngClass = max(LngtClass),
                   Number = sum(Number)/uniqLngClass)

dataSum %>% ggplot() + geom_line(aes(Year, uniqLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, uniqSpec)) + theme_bw()

dataSum %>% ggplot() + geom_line(aes(Year, minLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, maxLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, Number)) + theme_bw()

data_biomass %>% group_by(Year) %>% mutate(biomass = sum(Biomass)) %>% ggplot() + geom_line(aes(Year, biomass))

lob_size_spec %>% ggplot() + geom_boxplot(aes(Year, LngtClass, group = Year)) + theme_bw()
```

#### Create data bins

```{r data_bins}
# Create a key of what the min and max bin lengths and their weights are
species_splits <-data_biomass %>% 
  split(.$SpecCode)
# Get a key for each species and size class
# assumes 1cm bins for all species as written
data_bin_key <- species_splits %>% 
  map_dfr(function(species_df){
    
    # cod as tester
    #species_df <- species_splits$`Cod Atlantic`
    
    #pull the distinct length bins
    species_df <- species_df %>% 
      distinct(Year, LngtClass, .keep_all = T) %>% 
      arrange(LngtClass)
    
    
    # Add the max length for the bin, and its weight
    binned_df <- species_df %>% 
      mutate(
        #Set length max as 1cm larger than lower limit (assume all bins 1cm)
        LngtMax = LngtClass + 1, 
        wmax = exp(log(LWa) + LWb * log(LngtMax)) * 1000,
        wmax = ifelse(SpecCode %in% grams, wmax / 1000, wmax))  %>%
      select(Year, SpecCode, LWa, LWb, LngtMin = LngtClass, wmin = bodyMass, LngtMax, wmax,
             -c(Number, Biomass))

      
      
    # return the clean data
    return(binned_df)
    
    
    
  })

# Add the bins back into the original and clean up
dataBin <- data_biomass %>% 
  select(Year, SpecCode, LngtMin = LngtClass, Number, Biomass) %>% 
  left_join(data_bin_key, by = c("Year", "SpecCode", "LngtMin"))
```

#### Calculate size spectra slopes

```{r MLE_calcs}

# Map through instead of looping
mle_bins <- dataBin %>% 
  split(.$Year) %>% 
  imap_dfr(
    function(dataBinForLike, the_year){
    
    # Select the right columns
    dataBinForLike = dplyr::select(dataBinForLike,
                                   SpecCode,
                                   wmin,
                                   wmax,
                                   Number)
    
    # Set n, xmin, xmax
    n = sum(dataBinForLike$Number)
    xmin = min(dataBinForLike$wmin)
    xmax = max(dataBinForLike$wmax)
    
  
    
    # Get the likelihood calculation for the bins
    MLEbins.nSeaFung.oneyear.new  = calcLike(negLL.fn = negLL.PLB.bins.species,
                                             p = -1.9,
                                             suppress.warnings = TRUE,
                                             dataBinForLike = dataBinForLike,
                                             n = n,
                                             xmin = xmin,
                                             xmax = xmax,
                                             vecDiff = .9)
    # original vecDiff was 0.75
    
    # Store outputs in a dataframe
    MLEbins.nSeaFung.new = data.frame(Year = the_year,
                                      xmin = xmin,
                                      xmax = xmax,
                                      n = n,
                                      b = MLEbins.nSeaFung.oneyear.new$MLE,
                                      confMin = MLEbins.nSeaFung.oneyear.new$conf[1],
                                      confMax = MLEbins.nSeaFung.oneyear.new$conf[2])
    
    
    
    # Spit out the final output
    return(MLEbins.nSeaFung.new)
    
  }) 



# Need the standard error for weighted linear regression,
#  see eightMethods.count() for details:
MLEbins.nSeaFung.new = dplyr::as_tibble(mle_bins)
MLEbins.nSeaFung.new = dplyr::mutate(mle_bins,
                                     stdErr = (abs(confMin-b) +
                                               abs(confMax-b))/(2*1.96),
                                     Year = as.numeric(as.character(Year)))


MLEbins.nSeaFung.new %>% 
  ggplot(aes(Year, b)) +
  geom_segment(aes(x = Year, xend = Year, y = confMin, yend = confMax)) +
  geom_point(aes(y = b)) + 
  geom_smooth(formula = y ~ x) +
  labs(x = NULL,
       y = "Size Spectrum Slope (b)") + theme_bw()

```

#### Save outputs

```{r spring and fall}

# change name
mle_maine_fall <- MLEbins.nSeaFung.new
mle_maine_spring <- MLEbins.nSeaFung.new
mle_maine_both <- MLEbins.nSeaFung.new


mle_maine_fall$season <- "Fall"

mle_maine_spring$season <- "Spring"

mle_maine_both$season <- "Both"

mle_maine <- bind_rows(mle_maine_fall,mle_maine_spring, mle_maine_both)

write_csv(mle_maine, here::here("Indicators/MENH_size_spectra.csv"))

mle_maine %>% 
  ggplot(aes(Year, b)) +
  geom_segment(aes(x = Year, xend = Year, y = confMin, yend = confMax)) +
  geom_point(aes(y = b)) + 
  geom_smooth(formula = y ~ x) +
  facet_wrap(~season) +
  labs(x = NULL,
       y = "Size Spectrum Slope (b)") + theme_bw()
```

#### ME-NH Trawl Survey

Source: ME-NH Trawl Survey

Calculate stratified means

* Calculate catch per unit effort for each site for each year
* Multiply cpue by the depth strata area factor
* Group by stat area, sum the outputs 

```{r ME trawl survey data}
MEDMR_trawl <- read_csv(paste0(gmRi::box_path("Res_Data", "Maine_NH_Trawl"), "MaineDMR_Trawl_Survey_Catch_Data_2021-05-14.csv")) %>% 
  filter(Common_Name %in% lob_predators)


stat_area_trawl_key <- tibble("stat_area" = c(511, 512, 512, 513, 513),
                              "Region" = c(5,4,3,2,1))

DMR_strata_area <- tibble("Stratum" = c("1", "2", "3", "4"),
                          "1" = c(253.27, 214.22, 227.35, 225.65),
                          "2" = c(279.63, 191.23, 211.66, 263.49),
                          "3" = c(259.62, 262.90, 280.03, 183.69),
                          "4" = c(205.30, 206.12, 310.49, 170.72),
                          "5" = c(138.54, 220.49, 365.04, 196.11)) %>% 
  pivot_longer(cols = c("1", "2", "3", "4", "5"), names_to = "Region", values_to = "strata_area") %>% 
  group_by(Region) %>% 
  mutate(Stratum = as.numeric(Stratum),
         total = sum(strata_area),
         Region = as.numeric(Region)) %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum) %>% 
  summarise(strata_area = sum(strata_area),
            total = sum(total))

# average number of lobsters per tow 
u <- MEDMR_trawl %>% 
  left_join(stat_area_trawl_key) %>% 
  group_by(Season, Year, Tow_Number, stat_area, Stratum) %>% 
  summarise(n_lob_u = sum(Expanded_Weight_kg, na.rm = TRUE), .groups = "drop")

# average number of lobsters per trap haul at each depth stratum within a stat area
v <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>%  
  group_by(stat_area, Stratum, Year, Season) %>% 
  summarise(n_lob_v = mean(Expanded_Weight_kg, na.rm = TRUE), .groups = "drop")

# choose the relevant v which corresponds to the depth strata that u is in ("stat_area", "depth stratum", "Year"). Each site will have one w value
w <- left_join(v, u, by = c("stat_area", "Stratum", "Year", "Season")) %>% 
  mutate(w = (n_lob_v-n_lob_u)^2)

#sum of all w within the same depth strata in a stat area
x <- w %>% 
  group_by(Stratum, stat_area, Year, Season) %>% 
  summarise(x = sum(w), .groups = "drop")

#number of sites per depth stratum within a given stat area
y <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>% 
  group_by(stat_area, Stratum, Year, Season) %>% 
  summarise(Tow_number = unique(Tow_Number, na.rm = TRUE)) %>% 
  summarise(y = n(), .groups = "drop")

#if done correctly, each stat area per year should have three z values, one for each depth stratum
z <- left_join(x, y, by = c("stat_area", "Stratum", "Year", "Season")) %>% 
  mutate(z = x/(y-1))

# Calculate stat area variance 
MEDMR_vari <- left_join(z, DMR_strata_area, by = c("stat_area", "Stratum")) %>% 
  mutate(a = 1/(total^2),
         b = strata_area*(strata_area-y)*(z/y)) %>% 
  group_by(stat_area, Year, a, Season) %>% 
  summarise(stat_sum = sum(b), .groups = "drop") %>% 
  mutate(vari = a*stat_sum,
         sd = sqrt(vari))

medmr_vari_plot <- MEDMR_vari %>% 
  ggplot() + 
  geom_line(aes(Year, vari, col = as.factor(stat_area))) +
  facet_wrap(~Season)
```

#### Calculate cpue

```{r MEDMR_trawl cpue}

MEDMR_cpue <- MEDMR_trawl %>%
  left_join(stat_area_trawl_key) %>%  
  group_by(Year, Season, stat_area, Stratum)%>%
  mutate(tows=n_distinct(Tow_Number))%>%
  group_by(Year, Season, tows, stat_area, Stratum) %>%
  summarise(weight = sum(Expanded_Weight_kg,na.rm=TRUE), 
            catch=sum(Expanded_Catch,na.rm=TRUE))%>%
  mutate(weight_tow = weight/tows, 
         catch_tow = catch/tows) %>% 
  left_join(DMR_strata_area) %>% 
  mutate(stratified_wpue = weight_tow*(strata_area/total),
         stratified_cpue = catch_tow*(strata_area/total)) %>% 
  group_by(stat_area, Year, Season) %>% 
  summarise(lob_cpue = sum(stratified_cpue), 
            lob_index = sum(stratified_wpue),
            .groups = "drop") %>% 
  mutate(stat_area = as.factor(stat_area),
         name = "ME-NH_trawl")

MEDMR_cpue %>% 
  ggplot() +
  geom_line(aes(Year, lob_cpue, col = as.factor(stat_area))) +
  scale_color_discrete(name = "Stat area") +
  labs(y = "ME-NH Trawl predator cpue") +
  facet_wrap(~Season)


```

