---
title: "Lobster Indices"
author: "Matt Dzaugis"
date: "Updated on: `r Sys.Date()`"
output: 
  html_document:
    toc: TRUE
    toc_float:
        collapsed: FALSE
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message = F, 
                      warning = F, 
                      comment = "", 
                      fig.align = "center", 
                      fig.height = 6, 
                      fig.width = 6)

####  Packages  ####
library(sf)
library(tidyverse)
library(targets)
library(gmRi)
library(scales)

# paths
box_paths <- research_access_paths()
res_path <- box_paths$res

# Mapping polygons
new_england <- rnaturalearth::ne_states("united states of america", returnclass = "sf")
canada <- rnaturalearth::ne_states("canada", returnclass = "sf") 

# Set ggplot theme for figures
theme_set(theme_bw())

# Set theme up for maps
map_theme <- list(
  theme(
    panel.border       = element_rect(color = "black", fill = "transparent"),
    plot.background    = element_rect(color = "transparent", fill = "transparent"),
    line               = element_blank(),
    axis.title.x       = element_blank(), # turn off titles
    axis.title.y       = element_blank(),
    legend.position    = "bottom", 
    legend.title.align = 0.5)) 
```


`r use_gmri_style_rmd(css_file = "gmri_rmarkdown.css")`

# Lobster Predator Indices

The idea behind the lobster predator index is to use characteristics of the fish community that prey on lobster as a biological indicator for predicting patterns in lobster abundance/biomass/recruitment.


## Data Prep

The predator data used is sourced from the northeast trawl survey, conducted by the Northeast Fisheries Science Center. This survey runs every spring and fall sampling the seafloor using trawling gear towed behind a research vessel. Data from the survey includes abundance and size information for fishes found on the Northeastern US shelf.


### Trawl Survey Data Prep

Data from the survey is loaded for abundance-based analyses, with some filtering done to remove stratum that are sampled less-frequently following a survey design change in 2008. Following the standard tidy-up steps, each species then has their expected length-weight biomasses estimated using published growth curve equations.

```{r}
# Load clean trawl data, add length weight data
nefsc_clean <- load("/Users/mdzaugis/Box/RES_Data/NMFS_trawl/SURVDAT_current/NEFSC_BTS_all_seasons_03032021.RData")
nefsc_clean <- gmRi::gmri_survdat_prep(survdat = survey$survdat)
#nefsc_clean <- gmRi::gmri_survdat_prep(survdat_source = "most recent")
nefsc_clean <- add_lw_info(nefsc_clean, cutoff = FALSE)
```


### Subsetting Data to the Gulf of Maine

For the predator indices we are narrowing our attention to patterns within the Gulf of Maine. This is done to relate these patterns in predators to lobster observations in the area. These include regional landings, juvenile surveys, and recruitment indices. To do this data is only kept if it is sampled from within trawl stratum that  we typically associate with the Gulf of Maine

**Loading Trawl Survey Strata:**

```{r}
# Load the strata
survey_strata <- read_sf(str_c(res_path, "Shapefiles/BottomTrawlStrata/BTS_Strata.shp"))  %>% 
  janitor::clean_names() %>% 
  filter(strata >= 01010 ,
         strata <= 01760,
         strata != 1310,
         strata != 1320,
         strata != 1330,
         strata != 1350,
         strata != 1410,
         strata != 1420,
         strata != 1490) 


# Key to which strata = which regions
strata_key <- list(
  "Georges Bank"          = as.character(13:23),
  "Gulf of Maine"         = as.character(24:40),
  "Southern New England"  = str_pad(as.character(1:12), width = 2, pad = "0", side = "left"),
  "Mid-Atlantic Bight"    = as.character(61:76))

# Assign Areas by Strata
survey_strata <- survey_strata %>% 
  mutate(
    strata = str_pad(strata, width = 5, pad = "0", side = "left"),
    strata_num = str_sub(strata, 3, 4),
    area = case_when(
      strata_num %in% strata_key$`Georges Bank` ~ "Georges Bank",
      strata_num %in% strata_key$`Gulf of Maine` ~ "Gulf of Maine",
      strata_num %in% strata_key$`Southern New England` ~ "Southern New England",
      strata_num %in% strata_key$`Mid-Atlantic Bight` ~ "Mid-Atlantic Bight",
    TRUE ~ "Outside Major Study Areas")) %>% 
  select(finstr_id, strata, strata_num, area, a2, str2, set, stratuma, str3, geometry)
```

**Mapping the Trawl Regions:**

```{r}
# Map it against the coastline
ggplot() +
  geom_sf(data = new_england, size = 0.3) +
  geom_sf(data = canada, size = 0.3) +
  geom_sf(data = survey_strata, aes(fill = area)) +
  coord_sf(xlim = c(-77, -64.5), ylim = c(34.4, 45.75), expand = FALSE) +
  guides(fill = guide_legend(nrow = 2)) +
  theme_bw() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  map_theme +
  labs(subtitle = "Trawl Survey Regions")

```

**Subset to Gulf of Maine:**

```{r}
nefsc_gom <- nefsc_clean %>% 
  filter(survey_area == "GoM")
```


### Assigning Lobster Strata

Unfortunately the trawl survey and the surveys for lobster abundances are done using a different area stratification. To perform a more direct comparison among areas we reassign the trawl stations based on where they fall within the lobster survey strata.

#### Loading Strata

```{r}
# Lobster Strata
lobstrata <- read_sf(paste0(res_path, "Shapefiles/Statistical_Areas/Statistical_Areas_2010_withNames.shp"))
lobstrata <- rename_all(lobstrata, tolower)

# Subset the ones in GOM to plot faster
lobstrata_gom <- filter(lobstrata, id %in% c(464:467, 511:515, 521, 522, 526, 551, 561))


# Map against the Coast
ggplot() +
  geom_sf(data = new_england, size = 0.3) +
  geom_sf(data = canada, size = 0.3) +
  geom_sf(data = lobstrata_gom, fill = "transparent", show.legend = FALSE) +
  coord_sf(xlim = c(-71.5, -64.5), ylim = c(41.5, 45.75), expand = FALSE) +
  map_theme +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(subtitle = "Lobster Strata in the Gulf of Maine")
```

#### Strata Re-assignment

The reassignment is done by overlaying the starting locations for the trawl stations with the survey strata. Each station is assigned based on which strata is falls within. From there we can estimated area-weighted catch rates within each of them.

**Strata Re-assignment function:**

```{r}
# Assign statistical zone from new sf for stat zones
assign_stat_zones <- function(survdat, zone_sf, strata_col_in = "id", strata_col_out = "stat_zone", keep_NA = FALSE){
  
  # Transfer to shorthand names
  x <- as.data.frame(survdat)
  out_name_sym <- sym(strata_col_out)
  
  # use only station data for overlay/intersection
  stations <- distinct(x, cruise6, stratum, station, decdeg_beglat, decdeg_beglon)
  
  # Convert stations to sf
  stations_sf <- st_as_sf(stations, coords = c("decdeg_beglon", "decdeg_beglat"), crs = 4326, remove = FALSE)
  
  # Project to Lambert Conformal Conic
  lcc <- st_crs("+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-72 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ") 
  stations_sf <- st_transform(stations_sf, crs = lcc)
  
  # Prepare statistical zones in same CRS
  stratum <- st_transform(lobstrata, crs = lcc)
  
  # rename stratum column to match desired label
  names(stratum)[which(names(stratum) == strata_col_in)] <- strata_col_out
  
  # Identify points within each polygon/strata
  stations_sf <- st_join(stations_sf, stratum, join = st_within, left = TRUE)
  
  # Don't need to convert back b/c we kept coordinates
  stations_wgs <- st_drop_geometry(stations_sf)
  
  # Keep NA's or not?
  if(keep_NA == F){ stations_wgs <- filter(stations_wgs, is.na({{out_name_sym}}) == FALSE)}
  
  # Join station assignments back into full data
  out <- right_join(stations_wgs, x, by = c('cruise6', 'stratum', 'station', "decdeg_beglat", "decdeg_beglon")) %>% 
    mutate({{out_name_sym}} := as.character({{out_name_sym}}))

  # return the table
  return(out)
  
  }

```

**Assigning Lobster Strata:**

```{r}
# Assign those zones!
nefsc_lobsta_zones <- assign_stat_zones(survdat = nefsc_gom, 
                                        zone_sf = select(lobstrata, id, short_name, geometry), 
                                        strata_col_in = "id", 
                                        strata_col_out = "lobster_strata",
                                        keep_NA = FALSE)
```

**Validate by Mapping:**

```{r}
# make sf to check
gom_dat_sf <- nefsc_lobsta_zones %>% 
  distinct(decdeg_beglon, decdeg_beglat, lobster_strata) %>% 
  st_as_sf(coords = c("decdeg_beglon", "decdeg_beglat"), crs = 4326, remove = FALSE) 

# map check
ggplot() +
  geom_sf(data = gom_dat_sf, aes(color = lobster_strata)) +
  geom_sf(data = lobstrata, fill = "transparent") +
  geom_sf(data = new_england, size = 0.3) +
  geom_sf(data = canada, size = 0.3) +
  coord_sf(xlim = c(-71, -65.8), ylim = c(41, 44.5)) +
  map_theme +
  theme(legend.position = "bottom") +
  guides(color = guide_legend("Lobster Strata", title.position = "top", title.hjust = 0.5))
```

####  Reducing to Select Strata

For this analysis we want three main strata, and one aggregate of the three.
 - 511   
 - 512   
 - 513   
 
 These align with our other nearshore-focused indicators.

**Drop Strata with Incomplete Representation:**


```{r}
gom_dat <- nefsc_lobsta_zones %>% 
  filter(survey_area == "GoM",
         lobster_strata %in% c(511:513),
         #lobster_strata %not in% c(561, 522, 551, 526, 466, 467),
         is.na(lobster_strata) == FALSE)
```

**Validate by Mapping:**

```{r}
# make sf to check
gom_dat_sf_2 <- gom_dat %>% 
  distinct(decdeg_beglon, decdeg_beglat, lobster_strata) %>% 
  st_as_sf(coords = c("decdeg_beglon", "decdeg_beglat"), crs = 4326, remove = FALSE) 

# map check
ggplot() +
  geom_sf(data = new_england, size = 0.3) +
  geom_sf(data = canada, size = 0.3) +
  geom_sf(data = gom_dat_sf_2, aes(color = lobster_strata), show.legend = FALSE) +
  ggforce::geom_mark_ellipse(data = gom_dat_sf_2, 
                             aes(x = decdeg_beglon, y = decdeg_beglat, 
                                 group = lobster_strata, label = lobster_strata), 
                             color = "transparent", label.fill = "transparent") +
  coord_sf(xlim = c(-71, -65.8), ylim = c(41, 44.5)) +
  map_theme +
  theme(legend.position = "bottom") +
  guides(color = guide_legend("Lobster Strata", title.position = "top", title.hjust = 0.5))
```


### Calculate the Areas of Strata

To weight the catch rates of each stratum against one-another, they are weighted using their interior areas in square kilometers.

```{r}
#### Get Stratum Area in km2  ####

# Lambert Conformal Conic
lcc <- st_crs("+proj=lcc +lat_1=20 +lat_2=60 +lat_0=40 +lon_0=-72 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0 ")
lobstrata_lcc <- st_transform(lobstrata, lcc)

# Get areas
# rename the id column to match nefsc_lobsta
# drop geometry
zone_areas <- lobstrata_lcc %>% 
  rename(lobster_strata = id) %>% 
  mutate(lobster_strata = as.character(lobster_strata),
         area_m2 = st_area(lobstrata_lcc),
         area_km2 = units::set_units(area_m2, "km^2"),
         area_km2 = as.numeric(area_km2)) %>% 
  select(lobster_strata, area_km2) %>% 
  st_drop_geometry()


# merge in the areas
gom_weights <- left_join(gom_dat, zone_areas, by = "lobster_strata")


# Bar plot for area of each
gom_weights %>% 
  bind_rows(gom_weights %>% mutate(lobster_strata = "All Strata", area_km2 = sum(unique(gom_weights$area_km2)))) %>% 
  distinct(lobster_strata, area_km2) %>% 
  ggplot(aes(x = area_km2, fct_reorder(lobster_strata, as.numeric(area_km2)), fill = lobster_strata)) +
  geom_col() +
  scale_x_continuous(labels = comma_format()) +
  labs(y = "Lobster Strata", x = "Area (km^2)", fill = "Lobster Strata")
```


### Estimate Area-Stratified Catch

Area-stratified catch estimates are an approach that uses strata specific catch rates which are weighted by their overall size. This should be done before any species are filtered to make sure stations are not dropped by filtering out to a subset of species. At this point all the stations are in the data, and we need them to get effort correct.

Since we already reduced the potential area to these three strata, the total station count within a year is only for our area of interest. This will let us weight the three strata properly for getting the catch rates within the aggregate stratum.

**Area Stratification Function:**

```{r}
# added some tweaks to do catch rates in km2, and expand from there
stratify_lobster_strata_catch <- function(survdat_weights, area_label, area_col){
  
  # https://noaa-edab.github.io/survdat/articles/calc_strat_mean.html
  
 # # Testing:
 #  area_label <- "lobster_strata"
 #  area_col <- "area_km2"
 
  # column symbols from strings
  label_sym <- sym(area_label)
  area_km_sym <- sym(area_col)
  
  ####  1. Set Constants: 
  # Area covered by an albatross standard tow in km2
  alb_tow_km2 <- 0.0384
  # catchability coefficient - ideally should change for species guilds
  q <- 1
  
  
  ####  2. Stratum Area & Effort Ratios 
  # Get Annual Stratum Effort, and Area Ratios
  # The number of tows in each stratum by year
  # area of a stratum relative to total area of all stratum sampled that year
  
  # Get Total area of all strata sampled in each year (one each)
  total_stratum_areas <- group_by(survdat_weights, est_year)
  total_stratum_areas <- distinct(total_stratum_areas, {{label_sym}}, .keep_all = T)
  total_stratum_areas <- summarise(total_stratum_areas,
                                   tot_s_area =  sum({{area_km_sym}}, na.rm = T),
                                   .groups = "drop")
  
  
  # Calculate strata area relative to total area i.e. stratio or stratum weights
  survdat_weights <- left_join(survdat_weights, total_stratum_areas, by = "est_year")
  survdat_weights <- mutate(survdat_weights, 
                                   st_ratio = {{area_km_sym}} / tot_s_area,
                                   st_ratio = as.numeric(st_ratio))
  
  
  # We have total areas, now we want effort within each
  # Number of unique tows per stratum, within each season
  yr_strat_effort <- group_by(survdat_weights, est_year, season, {{label_sym}})
  yr_strat_effort <- summarise(yr_strat_effort, strat_ntows = n_distinct(id), .groups = "drop")
  
  # Add those yearly effort counts back for later
  # (area stratified abundance)
  survdat_weights <- left_join(survdat_weights, yr_strat_effort, 
                               by = c("est_year", "season", area_label))
  
  ####  4. Derived Stratum Area Estimates
  # All of the following are done based on size classes
  
  ## A. Catch per tow
  survdat_weights <- survdat_weights %>% 
    mutate(
      
      # Length Based :
      # Catch / tow, for that year & season group
      abund_tow_s   = numlen_adj / strat_ntows,
      # length based: biomass / tow
      lwbio_tow_s = biomass_g/1000 / strat_ntows,
      
      # Not Length Based: 
      biom_per_lclass = (biomass_g/1000 / n_len_class),
      biom_tow_s = biom_per_lclass / strat_ntows)
  
  
  ## B. Catch per km
  survdat_weights <- survdat_weights %>% 
    mutate(
      # Number / km2
      `abundance per km2` = abund_tow_s / alb_tow_km2,
      # kg / km2
      `kg per km2` = `biom_tow_s` / alb_tow_km2,
  )
  
  
  ## C. Stratified Total Abundance/Biomass
  survdat_weights <- survdat_weights %>% 
    mutate(
      ####  Area Extrapolations
      # Number of Individuals (Expected for entire stratum, based on density)
      `strat total abund` = `abundance per km2` * {{area_km_sym}},
      # Sum(bodymass) of Individuals (Expected for entire stratum, based on density)
      `strat total biomass (kg)` = `kg per km2` * {{area_km_sym}}
 )
  
  
  #biomass Variance - need to account for zero catch
  survdat_weights <- survdat_weights %>% 
    mutate(n_zero = strat_ntows - length(biomass_g),
           zero_var_b = n_zero * (0-biom_tow_s)^2,
           vari_b = (biomass_g/1000 - biom_tow_s)^2)
  
  survdat_weights <- survdat_weights %>% 
    group_by(cruise6, station, stratum, svspp, catchsex) %>% 
    mutate(sh_2_b = (zero_var_b + sum(vari_b))/(strat_ntows) - 1,
           sh_2_b = if_else(is.nan(sh_2_b), 0, sh_2_b)) %>% 
    ungroup()
  
  #abundance Variance - need to account for zero catch
  survdat_weights <- survdat_weights %>% 
    mutate(n_zero = strat_ntows - length(abundance),
           zero_var_a = n_zero * (0-abund_tow_s)^2,
           vari_a = (abundance - abund_tow_s)^2)
  
  survdat_weights <- survdat_weights %>% 
    group_by(cruise6, station, stratum, svspp, catchsex) %>% 
    mutate(sh_2_a = (zero_var_a + sum(vari_a))/(strat_ntows) - 1,
           sh_2_a = if_else(is.nan(sh_2_a), 0, sh_2_a)) %>% 
    ungroup()
  

  #### This section is for weighting stratified means, don't need here
  # ## D. Area weighted catch rates
  # # (catch rate, weighted by relative size of stratum)
  # # stratum area : area of all stratum sampled that season
  # survdat_weights <-  survdat_weights %>% 
  #   mutate(
  #     
  #     # Length Based:
  #     # Stratified mean abundance CPUE
  #     strat_mean_abund_s = abund_tow_s * st_ratio,
  #     # Stratified mean LW Biomass
  #     strat_mean_lwbio_s = lwbio_tow_s * st_ratio,
  #     
  #     # Not length based:
  #     # Stratified mean BIOMASS CPUE
  #     strat_mean_biom_s = biom_tow_s * st_ratio)
  # 
  # 
  # # Extrapolated Catch Totals from area-weighted rates
  # # convert from catch rate by area swept to total catch for entire stratum
  # survdat_weights <-  survdat_weights %>% 
  #   mutate(
  #     # Length Based:
  #     # Total Abundance
  #     strat_total_abund_s = round((strat_mean_abund_s * tot_s_area / alb_tow_km2) / q),
  #   
  #     # # Two options for to estimate lw biomass | Result is the same 4/20/2021
  #     #
  #     # Option 1: Individual LW Biomass * expanded abundance at length
  #     strat_total_lwbio_s = (ind_weight_kg * strat_total_abund_s) / q,
  #     
  #     # # Option 2: Size specific lw biomass / tow, expanded to total area
  #     # strat_total_lwbio_s  = (strat_mean_lwbio_s * tot_s_area / alb_tow_km2) / q
  #     
  #     # Not Length Based:
  #     # Total BIOMASS from the biomass of all lengths
  #     strat_total_biom_s = (strat_mean_biom_s * tot_s_area / alb_tow_km2) / q)
  
  
}

```


**Run Stratification:**

```{r}
# Run stratification
gom_strat <- stratify_lobster_strata_catch(survdat_weights = gom_weights, 
                                           area_label = "lobster_strata", 
                                           area_col = "area_km2")
# Tidy up?
# there are now two different "stratum columns" floating around
gom_strat <- gom_strat %>% 
  select(-c(strat_num, stratum, nafodiv, full_name)) 

```


### Pull Predator Complex

The last step before estimating any of the indices is to pull out the species that prey on lobster. There are 19 species included in this list:

```{r}
# List of species
lobster_predators <- c(
  "american lobster"
)

lobster_predators <- c(
  "atlantic halibut",
  "atlantic wolffish",
  "barndoor skate",
  "black sea bass",
  "atlantic cod",
  "fourspot flounder",
  "haddock",
  "little skate",
  "longhorn sculpin",
  "ocean pout",
  "red hake",
  "sea raven",
  "silver hake",
  "smooth skate",
  "spiny dogfish",
  "spotted hake",
  "thorny skate",
  "white hake",
  "winter flounder"
)
```


```{r}
library(sizeSpectra)

coeffs <- read_csv("/Users/mdzaugis/Box/Mills Lab/Projects/Lobster/indicators/Indicators/listfishusingfishbase.csv") %>% 
  janitor::clean_names() %>% dplyr::select(-x6)
# Filter
gom_predators <- gom_strat %>% 
  filter(comname %in% lobster_predators)

# Make another copy with both seasons included
gom_predators <- gom_predators %>% 
  bind_rows(gom_predators %>% mutate(season = "Both"))
```


```{r weight, echo=FALSE}
# Species that are in grams
grams <- c(
  "Monkfish",
  "Lumpfish",
  "Squid Short-Finned",
  "Sculpin Shorthorn",
  "Crab Jonah",
  "Wrymouth",
  "Fourbeard Rockling",
  "Crab Atlantic Rock",
  "Blenny Snake",
  "Smelt Rainbow",
  "Sand Lance American",
  "Menhaden Atlantic",
  "Silverside Atlantic",
  "Alligatorfish"
)

#### FILTER FOR SEASON

catch <- gom_predators %>% 
  filter(!is.na(ln_a))  %>% 
  mutate(ln_weight = (ln_a + b * log(length_cm)),
         weight = exp(ln_weight),
         freq_weight = weight*biom_per_lclass,
         a = exp(ln_a)) %>% 
  filter(season == "Both") %>%  #########33 LOOOK at season
  rename("Common_Name" = comname)

catch %>% 
  ggplot( aes( y = fct_reorder(Common_Name, weight, .fun = mean, .desc = TRUE), x = weight)) + 
    geom_boxplot() + 
    labs(x = "Weight (kg)", y = "Common Name")
```



```{r}
stat_area_key <- tibble("stat_area" = c(511,512,513),
                            "area" = c(296, 1381, 851))
stat_area_key$stat_area <- as.character(stat_area_key$stat_area)

cleanup_catch <- catch %>%
  select(est_year, lobster_strata, n_len_class, "Number" = numlen, "SpecCode" = Common_Name, "LngtClass" = length_cm, "LWa" = a, "LWb" = b, "bodyMass" = ind_weight_kg, "CPU_bio_per_hour" = freq_weight, "Year" = est_year) %>%
  distinct() %>% 
  left_join(stat_area_key, by = c("lobster_strata" = "stat_area"))

# sum numlen across strata

trawl_size_spec <- cleanup_catch  %>% 
  mutate(total_area = sum(unique(area))) %>% 
  group_by(Year, SpecCode, LngtClass)  %>% 
  summarise(
    Number = weighted.mean(Number, wt = area/total_area), 
    LWa = unique(LWa),
    LWb = unique(LWb),
    bodyMass = mean(bodyMass), .groups = "drop") 

  #group_by(Year, Region, Stratum, Common_Name, Length) %>% 
  #add_tally() %>%
  #summarise(Frequency = sum(Frequency)/n, .groups = "drop") %>%
  #left_join(., MENH_area, by = c("Region", "Stratum")) %>% 
  #group_by(Year, Season, Common_Name, Length) %>%
  #summarise(Frequency = weighted.mean(Frequency, wt, na.rm=TRUE), .groups = "drop") %>%
  #left_join(., coeffs, by = "Common_Name")


# Set bodymass lower limit
mass_cutoff <- .4 #kg

# Filter for lower end of gear selectivity
trawl_size_spec <- filter(trawl_size_spec, bodyMass >= mass_cutoff)


# Total number of fish
total_fish <- sum(trawl_size_spec$Number)

# Biomass data
data_biomass <- trawl_size_spec %>% 
  mutate(Biomass = Number * bodyMass)


```

Look to see if there is a big jump in smaller fish - is it something easy to point to - 

like lots of smaller fish and it is mostly x and y - take years and sort by size - histogram - see if something is dominating - 

tag team - kathy frame it up and ends with future projections 

Pulled out size spectra and maine cc b/c they are new

look into size spec

```{r look at data}
dataSum <- trawl_size_spec %>% group_by(Year) %>% 
  dplyr::summarise(uniqLngClass = length(unique(LngtClass)),
                   uniqSpec = length(unique(SpecCode)),
                   minLngClass = min(LngtClass),
                   maxLngClass = max(LngtClass),
                   Number = sum(Number)/uniqLngClass)

dataSum %>% ggplot() + geom_line(aes(Year, uniqLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, uniqSpec)) + theme_bw()

dataSum %>% ggplot() + geom_line(aes(Year, minLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, maxLngClass)) + theme_bw()
dataSum %>% ggplot() + geom_line(aes(Year, Number)) + theme_bw()

data_biomass %>% group_by(Year) %>% mutate(biomass = sum(Biomass)) %>% ggplot() + geom_line(aes(Year, biomass))

trawl_size_spec %>% ggplot() + geom_boxplot(aes(Year, LngtClass, group = Year)) + theme_bw()
```



```{r}
# Create a key of what the min and max bin lengths and their weights are
species_splits <-data_biomass %>% 
  split(.$SpecCode)
# Get a key for each species and size class
# assumes 1cm bins for all species as written
data_bin_key <- species_splits %>% 
  map_dfr(function(species_df){
    
    # cod as tester
    #species_df <- species_splits$`Cod Atlantic`
    
    #pull the distinct length bins
    species_df <- species_df %>% 
      distinct(Year, LngtClass, .keep_all = T) %>% 
      arrange(LngtClass)
    
    
    # Add the max length for the bin, and its weight
    binned_df <- species_df %>% 
      mutate(
        #Set length max as 1cm larger than lower limit (assume all bins 1cm)
        LngtMax = LngtClass + 1, 
        wmax = exp(log(LWa) + LWb * log(LngtMax)) * 1000)  %>%
      select(Year, SpecCode, LWa, LWb, LngtMin = LngtClass, wmin = bodyMass, LngtMax, wmax,
             -c(Number, Biomass))

      
      
    # return the clean data
    return(binned_df)
    
    
    
  })

# Add the bins back into the original and clean up
dataBin <- data_biomass %>% 
  select(Year, SpecCode, LngtMin = LngtClass, Number, Biomass) %>% 
  left_join(data_bin_key, by = c("Year", "SpecCode", "LngtMin"))
```



```{r MLE_calcs}

# Map through instead of looping
mle_bins <- dataBin %>% 
  split(.$Year) %>% 
  imap_dfr(
    function(dataBinForLike, the_year){
    
    # Select the right columns
    dataBinForLike = dplyr::select(dataBinForLike,
                                   SpecCode,
                                   wmin,
                                   wmax,
                                   Number)
    
    # Set n, xmin, xmax
    n = sum(dataBinForLike$Number)
    xmin = min(dataBinForLike$wmin)
    xmax = max(dataBinForLike$wmax)
    
  
    
    # Get the likelihood calculation for the bins
    MLEbins.nSeaFung.oneyear.new  = calcLike(negLL.fn = negLL.PLB.bins.species,
                                             p = -1.9,
                                             suppress.warnings = TRUE,
                                             dataBinForLike = dataBinForLike,
                                             n = n,
                                             xmin = xmin,
                                             xmax = xmax,
                                             vecDiff = .9)
    # original vecDiff was 0.75
    
    # Store outputs in a dataframe
    MLEbins.nSeaFung.new = data.frame(Year = the_year,
                                      xmin = xmin,
                                      xmax = xmax,
                                      n = n,
                                      b = MLEbins.nSeaFung.oneyear.new$MLE,
                                      confMin = MLEbins.nSeaFung.oneyear.new$conf[1],
                                      confMax = MLEbins.nSeaFung.oneyear.new$conf[2])
    
    
    
    # Spit out the final output
    return(MLEbins.nSeaFung.new)
    
  }) 



# Need the standard error for weighted linear regression,
#  see eightMethods.count() for details:
MLEbins.nSeaFung.new = dplyr::as_tibble(mle_bins)
MLEbins.nSeaFung.new = dplyr::mutate(mle_bins,
                                     stdErr = (abs(confMin-b) +
                                               abs(confMax-b))/(2*1.96),
                                     Year = as.numeric(as.character(Year)))


MLEbins.nSeaFung.new %>% 
  ggplot(aes(Year, b)) +
  geom_segment(aes(x = Year, xend = Year, y = confMin, yend = confMax)) +
  geom_point(aes(y = b)) + 
  geom_smooth(formula = y ~ x) +
  labs(x = NULL,
       y = "Size Spectrum Slope (b)") + theme_bw()

```



```{r spring and fall}

# change name
mle_maine_fall <- MLEbins.nSeaFung.new
mle_maine_spring <- MLEbins.nSeaFung.new
mle_maine_both <- MLEbins.nSeaFung.new


mle_maine_fall$season <- "Fall"

mle_maine_spring$season <- "Spring"

mle_maine_both$season <- "Both"

mle_maine <- bind_rows(mle_maine_fall,mle_maine_spring, mle_maine_both)

write_csv(mle_maine, here::here("Indicators/NEFSC_size_spectra.csv"))

mle_maine %>% 
  ggplot(aes(Year, b)) +
  geom_segment(aes(x = Year, xend = Year, y = confMin, yend = confMax)) +
  geom_point(aes(y = b)) + 
  geom_smooth(formula = y ~ x) +
  facet_wrap(~season) +
  labs(x = NULL,
       y = "Size Spectrum Slope (b)") + theme_bw()
```

`r insert_gmri_footer()`


