---
title: "Update_buoys_erddap"
author: "Matt Dzaugis"
date: "3/2/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rerddap)
library(ncdf4)
library(tidyverse)
library(lubridate)
```

```{r UpdateBuoy Overview, warning=FALSE, message=FALSE, results='hide'}
BB<-UpdateBuoy(buoy="B01",smoothing=FALSE,surfbot=FALSE)
BE<-UpdateBuoy(buoy="E01",smoothing=FALSE,surfbot=FALSE)
BF<-UpdateBuoy(buoy="F01",smoothing=FALSE,surfbot=FALSE)
BI<-UpdateBuoy(buoy="I01",smoothing=FALSE,surfbot=FALSE)
BM<-UpdateBuoy(buoy="M01",smoothing=FALSE,surfbot=FALSE)
BN<-UpdateBuoy(buoy="N01",smoothing=FALSE,surfbot=FALSE)
BA<-UpdateBuoy(buoy="A01",smoothing=FALSE,surfbot=FALSE)

Buoys <-list("Buoy_A" = BA, "Buoy_B" = BB, "Buoy_E" = BE, "Buoy_F" = BF, "Buoy_I" = BI, "Buoy_M" = BM, "Buoy_N" = BN)

Buoys_unlisted <- reshape2::melt(Buoys, id.vars = "Date") %>% 
  rename("Type" = variable, "Variable" = L3, "Depth" = L2, "name" = L1) %>% 
  mutate(Depth = as.double(parse_number(Depth)), 
         name = str_replace_all(name, "_", " "), 
         Type = paste("raw"),
         Date = as.Date(Date),
         dayz = day(Date),
         mon = month(Date),
         yr = year(Date)) %>% 
  group_by(Variable, Depth, name, dayz, mon) %>%
  mutate(clim = mean(value, na.rm = TRUE)) %>% ungroup() %>%
  mutate(Anomaly = value - clim) %>% dplyr::select(name, Date, Depth, Variable, Anomaly, "raw" = value) %>% 
  pivot_longer(., cols = c(Anomaly, raw), names_to = "Type", values_to = "Values")


data.table::fwrite(Buoys_unlisted, "Data/buoys.csv")
```

### loadNERACOOSnetcdf

This function opens a downloaded netcdf file of a standard format, `filename` and extracts the variable called in `varname`. This function is called within `UpdateBuoys` and does not need to be modified by the user.

```{r loadNERACOOSnetcdf, include=TRUE}
loadNERACOOSnetcdf<-function(filename,varname){
  
  # open the netcdf file, extract the variable, and save as a vector
  ncid<-nc_open(filename)
  V<-ncvar_get(ncid,varid= varname)
  V<-as.numeric(V)
  
  # look for any flagged values, replace with NaN
    # values >0 indicate values out of range, broken sensor, or invalid input 
  Q<-ncvar_get(ncid,varid=paste(varname,'_qc',sep=''))
  I<-which(Q>0)
  V[I]<-NaN
  
  # load the time values
  time<-ncvar_get(ncid,varid='time')
  depth <- ncvar_get(ncid, varid = "depth")
  
  # close the netcdf file and combine time and variable output into a dataframe
  nc_close(ncid)
  
  time <- as.POSIXct(time, origin = "1970-01-01T00:00:00Z")
  time <- data.frame(time)
  
  df <- data.frame(cbind(time, V, depth))
  
# Convert decimal day to Year-month-day, calculate daily means
  df <- df %>% mutate(day = day(time)) %>% 
    mutate(month = month(time)) %>% mutate(year = year(time))
  df <- df %>% group_by(year, month, day, depth) %>% summarise(daily_mean = mean(V, na.rm = TRUE)) %>%
    mutate(Date = as.Date(paste(year, month, day, sep = "-")))

# Create a tibble with all of the dates to make a full time series including missing values
  start_date <- head(as.Date(df$Date), n=1)
  end_date <- tail(as.Date(df$Date), n=1)
  ts_date <- tibble(Date = seq(as_date(start_date), as_date(end_date), by = "day"))
  df <- left_join(ts_date, df[,c("Date", "daily_mean", "depth")], by = "Date", all.x =TRUE)  
}
```


### surfbotcorr

This function calculates the surface to bottom linear regression to fill in gaps in the 50m temperature data, if the surface temperature is available. `surfbotcorr` is called with a logical `TRUE` or `FALSE` by the user as an argument in the `UpdateBuoy` function.

```{r surfbotcorr, include=TRUE}
surfbotcorr<-function(buoydat){
  # extract the surface (1m) amd bottom (50m) temperature data from each buoy
  bottomT<-buoydat %>% filter(depth == 50)
  surfT<-buoydat %>% filter(depth == 1)
  
  # match rows with the same dates to make a table of 1m and 50m temperature data
  Table <- left_join(surfT, bottomT, by = "Date", all.x = TRUE)
  
  #create a column of ordinal days and year
  Table <- Table %>% mutate(yr = year(Date)) %>% mutate(ord = yday(Date))
  
  # Linear model: 50 meter temperature dependent on 1 meter temperature grouped by ordinal day
  lm1 <-  Table %>% group_by(ord) %>% do(broom::tidy(lm(daily_mean.y ~ daily_mean.x, data = .))) %>% 
    dplyr::select(ord, term, estimate) %>% spread(key = term, value = estimate) 
  
  # combine the temperature data with the slope and intercept of the lm
  Table <- merge(Table, lm1[,c("ord", "(Intercept)", "daily_mean.x")], by = "ord", all.x = TRUE)
  Table <- Table %>% group_by(ord) %>% mutate(sum = sum(is.na(daily_mean.y)))
  end <- length(Table$Date)
  
  # using the lm coefficients, calculate the 50m temperature based on the 1m temperature for each missing 50m value
  for(i in 1:end){
    if(is.na(Table$daily_mean.y[i]) & !is.na(Table$daily_mean.x.x[i]) & Table$sum[i] <10)
    {Table$daily_mean.y[i]<-Table$'(Intercept)'[i]+Table$daily_mean.x.y[i]*Table$daily_mean.x.x[i]}
  }
  Table <- ungroup(Table)
  Table <- Table %>% rename(daily_mean = daily_mean.y)
  bottomT <- Table %>% dplyr::select(Date, daily_mean, "depth" = depth.y) %>% bind_rows(surfT)
  return(bottomT)
} 
```

### library_check

This is a helper function that checks if the required packages are installed and loaded into the workspace. the only argument, `libraries` is a vector of required library names. 

```{r library_check, include=TRUE}
library_check<- function(libraries) {
  lapply(libraries, FUN = function(x) {
    if (!require(x, character.only = TRUE)) {
      install.packages(x, dependencies = TRUE)
      library(x, character.only = TRUE)
    }
  })
}
```



```{r update buoys}

UpdateBuoy<-function(buoy,smoothing,surfbot){
  
  # dplyr::select buoy: "B01", "E01", "F01", "I01", "M01", or "N01" named in argument 1
  buoy<- buoy
  
  # create the base path and file
  path<-"http://www.neracoos.org/erddap/tabledap/"
  file<-"_sbe37_all.nc?station%2Ctime%2Ctemperature%2Ctemperature_qc%2Csalinity%2Csalinity_qc%2Csigma_t%2Csigma_t_qc%2Clongitude%2Clatitude%2Cdepth"
  destfile <- paste("Data/Buoy_files/", buoy,"_",Sys.Date(), ".nc", sep="")
  
  download.file(paste(path, buoy, file, sep=""), destfile = destfile, mode="wb")

  filename <- paste(here::here(destfile))
  
# TEMPERATURE
  temp <- loadNERACOOSnetcdf(filename = filename, "temperature")

  temp <- temp %>% rename("temp" = daily_mean)
  
# SALINITY
  salt <- loadNERACOOSnetcdf(filename = filename, "salinity")
  salt <- salt %>% rename("salt" = daily_mean)
  
# DENSITY
  den <- loadNERACOOSnetcdf(filename = filename, "sigma_t")
  den <- den %>% rename("den" = daily_mean)
  
# To smooth or not to smooth
    if(smoothing==TRUE){
      
      # 8-day mean loop
      rolling_8 <- rollify(mean, window = 8)
      temp <- temp %>% group_by(depth) %>% mutate(rolling_mean = rolling_8(daily_mean)) 
      salt <- salt %>% group_by(depth) %>% mutate(rolling_mean = rolling_8(daily_mean))
      den <- den %>% group_by(depth) %>% mutate(rolling_mean = rolling_8(daily_mean))
    }
  
  if(surfbot == TRUE){
    temp <- surfbotcorr(temp)
  }
  
  if(buoy %in% c("A01", "B01", "E01", "F01", "I01")){
   depth_1m <- list("temp" = dplyr::select(filter(temp, depth == 1), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 1), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 1), Date, "daily_mean" = den))
   
   depth_20m <- list("temp" = dplyr::select(filter(temp, depth == 20), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 20), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 20), Date, "daily_mean" = den))
   
   depth_50m <- list("temp" = dplyr::select(filter(temp, depth == 50), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 50), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 50), Date, "daily_mean" = den)) 
   B <- list(depth_1m=depth_1m, depth_20m=depth_20m, depth_50m=depth_50m)}

 if(buoy == "M01"){
   depth_1m <- list("temp" = dplyr::select(filter(temp, depth == 1), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 1), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 1), Date, "daily_mean" = den))
   
   depth_20m <- list("temp" = dplyr::select(filter(temp, depth == 20), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 20), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 20), Date, "daily_mean" = den))
   
   depth_50m <- list("temp" = dplyr::select(filter(temp, depth == 50), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 50), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 50), Date, "daily_mean" = den))
   
    depth_100m <- list("temp" = dplyr::select(filter(temp, depth == 100), Date, "daily_mean" = temp), 
                  "sal" = dplyr::select(filter(salt, depth == 100), Date, "daily_mean" = salt),
                  "density" = dplyr::select(filter(den, depth == 100), Date, "daily_mean" = den))

   depth_150m <- list("temp" = dplyr::select(filter(temp, depth == 150), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 150), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 150), Date, "daily_mean" = den))
   
   depth_200m <- list("temp" = dplyr::select(filter(temp, depth == 200), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 200), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 200), Date, "daily_mean" = den)) 
   
   depth_250m <- list("temp" = dplyr::select(filter(temp, depth == 250), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 250), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 250), Date, "daily_mean" = den))
     
   B <- list(depth_1m=depth_1m, depth_20m=depth_20m, depth_50m=depth_50m, depth_100m=depth_100m, depth_150m=depth_150m, depth_200m=depth_200m, depth_250m=depth_250m)
 }
 
 
  if(buoy == "N01"){
    depth_1m <- list("temp" = dplyr::select(filter(temp, depth == 1), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 1), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 1), Date, "daily_mean" = den))
   
   depth_20m <- list("temp" = dplyr::select(filter(temp, depth == 20), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 20), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 20), Date, "daily_mean" = den))
   
   depth_50m <- list("temp" = dplyr::select(filter(temp, depth == 50), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 50), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 50), Date, "daily_mean" = den))
   
    depth_100m <- list("temp" = dplyr::select(filter(temp, depth == 100), Date, "daily_mean" = temp), 
                    "sal" = dplyr::select(filter(salt, depth == 100), Date, "daily_mean" = salt),
                    "density" = dplyr::select(filter(den, depth == 100), Date, "daily_mean" = den))
   
     depth_150m <- list("temp" = dplyr::select(filter(temp, depth == 150), Date, "daily_mean" = temp), 
                      "sal" = dplyr::select(filter(salt, depth == 150), Date, "daily_mean" = salt),
                      "density" = dplyr::select(filter(den, depth == 150), Date, "daily_mean" = den))
     
     depth_180m <- list("temp" = dplyr::select(filter(temp, depth == 180), Date, "daily_mean" = temp), 
                      "sal" = dplyr::select(filter(salt, depth == 180), Date, "daily_mean" = salt),
                      "density" = dplyr::select(filter(den, depth == 180), Date, "daily_mean" = den)) 
     B <- list(depth_1m=depth_1m, depth_20m=depth_20m, depth_50m=depth_50m, depth_100m=depth_100m, depth_150m=depth_150m, depth_180m=depth_180m)
   
 }
 
 return(B)
  
}
  
```







