---
title: "Update_buoys_erddap"
author: "Matt Dzaugis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rerddap)
library(ncdf4)
library(tidyverse)
library(lubridate)

source(here::here("Code/UpdateBuoy_functions.R"))
```

## UpdateBuoy overview

`UpdateBuoy(buoy, smoothing, surfbot)`

`UpdateBuoy` has four arguments: `buoy`, `path.location`, `smoothing`, and `surfbot`. There are no default choices for the arguments, so each argument has to be called in the function. This function takes a little over 1.5 min in total to download and extract data from all 6 buoys; so each buoy takes about 15 seconds.  

* `buoy` calls on one of six NERACOOS buoys: "B01", "E01", "F01", "I01", "M01", or "N01". 
* `smoothing` is logical, TRUE or FALSE. If TRUE, the data are smoothed with an 8 day rolling mean and included in the list as a separate vector with the daily mean and date. If FALSE the rolling mean is not calculated.
* `surfbot` is logical, TRUE or FALSE and only applies to temperature data. If TRUE, missing 50m temperature data are filled in using modeled data from a linear regression with surface data. If `smoothing = TRUE` AND `surfbot = TRUE`, then the 8 day rolling mean is not calculated for 50m temperature. 

`UpdateBuoy` uses packages `here`, `ncdf4`, `lubridate`, `tidyverse`, and `tibbletime`. The `library_check` function can be used to install or load these functions and is included at the end of this document. 

### Run the function

```{r UpdateBuoy Overview, warning=FALSE, message=FALSE, results='hide'}
BB<-UpdateBuoy(buoy="B01",smoothing=FALSE,surfbot=FALSE)
BE<-UpdateBuoy(buoy="E01",smoothing=FALSE,surfbot=FALSE)
BF<-UpdateBuoy(buoy="F01",smoothing=FALSE,surfbot=FALSE)
BI<-UpdateBuoy(buoy="I01",smoothing=FALSE,surfbot=FALSE)
BM<-UpdateBuoy(buoy="M01",smoothing=FALSE,surfbot=FALSE)
BN<-UpdateBuoy(buoy="N01",smoothing=FALSE,surfbot=FALSE)
BA<-UpdateBuoy(buoy="A01",smoothing=FALSE,surfbot=FALSE)

Buoys <-list("Buoy_A" = BA, "Buoy_B" = BB, "Buoy_E" = BE, "Buoy_F" = BF, "Buoy_I" = BI, "Buoy_M" = BM, "Buoy_N" = BN)

Buoys_unlisted <- reshape2::melt(Buoys, id.vars = "Date") %>% 
  rename("Type" = variable, "Variable" = L3, "Depth" = L2, "name" = L1) %>% 
  mutate(Depth = as.double(parse_number(Depth)), 
         name = str_replace_all(name, "_", " "), 
         Type = paste("raw"),
         Date = as.Date(Date),
         dayz = day(Date),
         mon = month(Date),
         yr = year(Date)) %>% 
  group_by(Variable, Depth, name, dayz, mon) %>%
  mutate(clim = mean(value, na.rm = TRUE)) %>% ungroup() %>%
  mutate(Anomaly = value - clim) %>% dplyr::select(name, Date, Depth, Variable, Anomaly, "raw" = value) %>% 
  pivot_longer(., cols = c(Anomaly, raw), names_to = "Type", values_to = "Values")


data.table::fwrite(Buoys_unlisted, here::here("Data/Buoy_data.csv"))
```






